Spatial integration of sensory input and motor output in Pseudomonas aeruginosa chemotaxis through colocalized distribution

Version 1

Abstract

Abstract

The opportunistic pathogen Pseudomonas aeruginosa serves as a model organism for studying multiple signal transduction pathways. The chemoreceptor cluster, a core component of the chemotaxis pathway, is assembled from hundreds of proteins. The unipolar distribution of receptor clusters has long been recognized, yet the precise mechanism governing their assembly remains elusive. Here, we directly observed the relative positions of the flagellar motor and chemoreceptor cluster using flagellar filament labeling and gene editing techniques. Surprisingly, we found that both are located at the same cell pole, with the distribution pattern controlled by the polar anchor protein FlhF. Additionally, the efficient assembly of the chemoreceptor cluster is partially dependent on the integrity of the motor structure. Furthermore, we discovered that overexpression of the chemotaxis regulatory protein CheY leads to high intracellular levels of the second messenger c-di-GMP, triggering cell aggregation. Therefore, the colocalization of the chemoreceptor cluster and flagellum in P. aeruginosa serves to avoid cross-pathway signaling interference, enabling cells to conduct various physiological activities in an orderly manner.

Introduction

In biological systems, the process by which proteins self-assemble into organized complex structures is ubiquitous1, 2. Pole-to-pole protein oscillations in the Min system ensure that the FtsZ ring, a crucial component of cell division, is placed precisely in the middle of the cell body3, 4. In response to antibiotic treatment and heat stress, some bacteria generate multiple protein foci throughout the cytoplasm5, 6. The polar localization of the secretion system, such as the type Ⅵ secretion system, is mediated by targeting proteins and potentially facilitates host-pathogen interactions7, 8. Thus, spontaneous assembly is accompanied by the formation of spatial patterns that order the intracellular environment.

To grow and colonize within the host, bacteria have evolved a mechanism for migrating towards more favorable environments. The locomotion organ (the flagellar motor) and the chemotaxis pathway play crucial roles in achieving this goal. The former provides physical drive, while the latter offers directional guidance. The initiation of cell reorientation is controlled by motor switching, which is governed by the chemotaxis two-component signaling pathway. This pathway comprises transmembrane chemoreceptors, a histidine kinase CheA, an adaptor protein CheW, a response regulator CheY, and two adaptation proteins, CheB and CheR. Receptors collaborate with CheA and CheW to form chemosensory arrays in a hexagonal pattern9, 10. Upon phosphorylation of CheY by the kinase at the array, it freely diffuses to the motor and binds to FliM, a component of the motor switch complex, to affect motor switching11–13. It is worth noting that there exists diversity among different bacterial species in the placement of flagellar motors14, whereas chemoreceptor clusters are typically found at one or both ends of the cell15. These two structural units are functionally connected. However, it remains unclear whether there is an interaction in their spatial distribution and what the specific regulatory mechanism might be.

Pseudomonas aeruginosa, a common human opportunistic pathogen, possesses four chemosensory pathways that perform distinct functions and are stimulated by signal binding to 26 chemoreceptors16. Among them, F6 pathway-related receptors that mediate chemotaxis, as well as the flagellum are localized at the same cell pole17, 18. Furthermore, the polar anchor protein FlhF has been reported to guide the flagellum to grow at the cell pole19, 20. However, the mechanism by which chemoreceptor array-related proteins aggregate and self-assemble at the cell pole is controversial. The distribution of the chemosensory arrays in the peritrichous bacterium Escherichia coli has been extensively studied and attributed to several mechanisms, including stochastic self-assembly21, 22, membrane curvature sorting23, 24, and inefficient clustering in the lateral region25. However, these mechanisms cannot explain the unipolar distribution pattern observed in P. aeruginosa. Unlike E. coli, specific genes responsible for the placement of the chemosensory array have been identified in several bacterial species. In Caulobacter crescentus, chemosensory arrays assemble at the new cell pole with the help of TipN and TipF proteins26. A tripartite ParC-ParP-CheA interaction network was reported to promote polar localization of chemosensory array in Vibrio parahaemolyticus27, 28. However, as there are no related genes in P. aeruginosa, the mechanism of its chemoreceptor array distribution remains a mystery.

Here, we combined gene editing and in vivo fluorescence imaging of flagellar filaments to directly observe the distribution of chemoreceptor arrays and flagellar motors, proposing a cooperative construction model of chemotaxis network and flagella during the entire division cycle of P. aeruginosa. The core focus of this study is to clarify the regulatory mechanism of its chemoreceptor array distribution. We found a substantial association between the assembly of the flagellar motor and the chemoreceptor array. The complete assembly of the motor serves as a partial prerequisite for the assembly of the receptor array, and its assembly site is also regulated by the polar anchor protein FlhF. Furthermore, by introducing exogenously expressed CheY protein, we found that this triggers the expression of c-di-GMP at a high level. From this, we infer that the colocalization of the chemoreceptor array and the flagellum in P. aeruginosa avoids the cross-pathway interference of signaling molecules, thus providing a guarantee for the coexistence of multiple chemosensory pathways.

Results

Robust generation of daughter cell with both chemotaxis network and flagellar motor

CheY has been shown to colocalize with chemoreceptors29. To visualize the distribution of chemoreceptor arrays in cells, we fused the gene encoding the enhanced yellow fluorescent protein (eYFP) to the cheY gene in the P. aeruginosa chromosome (Fig. 1A). All genetic modifications were carried out at the native position on the chromosome, allowing the expression of chemotaxis proteins in precise stoichiometry from their natural promoters (Fig. 1B). The mutant strain with cheY-eyfp was able to moderately expand on soft agar plates17, proving that its chemotaxis behavior was not notably affected, so it was selected for subsequent experiments. As shown in Fig. 1C, CheY-eYFP was mainly located at the cell pole, suggesting that CheA, CheW and CheY in P. aeruginosa form a signal transduction complex that was primarily distributed at the cell pole, as described previously17. From our measurements, nearly 90% (332/372) of cells contain obvious receptor clusters, and factors such as fluorescence bleaching may have caused the loss of fluorescent spots in some cells. In addition, we observed the presence of fluorescent spots at both ends of some cells with a large aspect ratio (probably approaching cell division), indicating that the newly generated progeny cells will have a complete chemotaxis network. Next, we sought to observe the distribution of the flagella in cells simultaneously to understand the motility of future progeny.

A. Schematic diagram of the chemotaxis signal transduction network and flagellar motor of P. aeruginosa. Fusion protein CheY-EYFP and fluorescently labeled flagellar filaments were used as markers to indicate the position of the chemosensory array and motor, respectively. B. The labeling mechanism of flagellar filaments and chemotaxis regulatory protein CheY. Filaments (with cysteine point mutation FliCT394C) were labeled through sulfhydryl-maleimide conjugation, and cheY-eyfp fusion with a 3× glycine linker was used to visualize chemosensory array positions. C. Subcellular localization of CheY-EYFP in the wild-type strain of P. aeruginosa. CheY-EYFP is mainly located at the single cell pole, and the white arrow points to individuals with obvious chemosensory arrays at both cell poles, which generally have a large aspect ratio of the cell body. The yellow dashed box marks the cell outline. D. The merged imaging of flagellar filaments and CheY-EYFP in the wild-type strain of P. aeruginosa, where flagellar motor and chemosensory array colocalize in cells. White arrows point to individuals about to be divided, and the yellow dashed box marks the cell outline. The scale bar is 1 μm.

Our recent study demonstrated the successful labeling of the flagellar filament with thiol-reactive fluorescent dye by introducing cysteine mutations into the flagellin FliC30, 31 (Fig. 1B). To avoid fluorescence interference, we utilized a dye with an excitation peak near 568 nm, considering that the excitation peak of eYFP is 513 nm, and the emission peak is 527 nm. We employed a xenon lamp as the excitation light source and switched filters to enable simultaneous fluorescence imaging of flagellar filaments and chemosensory arrays within the same cell. Remarkably, our findings revealed a surprising colocalization of chemosensory arrays and flagellar filaments at the same end of the cell body, and this colocalization was consistent, as shown in Fig. 1D. Similar to the chemoreceptor arrays, we observed the phenomenon of double flagella symbiosis in cells with a large aspect ratio, suggesting that P. aeruginosa has evolved a cell division mechanism with precise timing regulation. Before cell division, both poles of the mother cell assembled chemoreceptor arrays and flagellar motors. This ensures that daughter cells possess complete motility and chemotaxis, thereby greatly enhancing the environmental adaptability of the population.

FlhF mediates proper unipolar localization of the chemoreceptor array

Given that the physiological and physical environments of both cell poles are nearly identical, it is unlikely that the unipolar distribution of the chemoreceptor array can be attributed to passive regulatory factors. The number and distribution of flagellar motors and chemoreceptor arrays vary among different bacterial species14, 15, and the localization system for flagella has been well studied. Specifically, the FlhF-FlhG system, discovered in several monotrichous bacterial species, has been shown to control the location and number of flagella32, 33. In P. aeruginosa, a knockout of flhF leads to mis-localized flagellar assembly19, 20. Considering the consistent colocalization pattern between chemosensory arrays and flagellar motors in P. aeruginosa, we speculate that the distribution of receptor clusters is also regulated by similar molecular mechanisms.

To investigate the role of FlhF in the localization of receptor arrays, we constructed a ΔflhF strain for fluorescence observation. The results revealed that the chemoreceptor arrays no longer grow robustly at the cell pole (Fig. 2A), and the assembly positions of the flagellar motor change accordingly, colocalizing with the receptor arrays (Fig. 2B). We also quantified the proportion of individuals with obvious receptor clusters, which was more than 80% (182/221), similar to the wild-type strain. These experimental results suggest that the polar anchoring protein FlhF has a minimal impact on the assembly efficiency of P. aeruginosa’s chemosensory array, but it strictly controls its unipolar distribution. Furthermore, a consistent colocalization between the flagellar motor and chemosensory array was observed, independent of FlhF. Thus, it is crucial to determine whether there is a causal relationship in the assembly order of these two structural units.

A. Subcellular localization of CheY-EYFP in the ΔflhF strain of P. aeruginosa. CheY-EYFP is no longer robustly distributed at the single cell pole. The yellow dashed box marks the cell outline. B. The merged imaging of flagellar filaments and CheY-EYFP in the ΔflhF strain of P. aeruginosa, flagellar motor and chemosensory array still colocalize in cells. The yellow dashed box marks the cell outline. The scale bar is 1 μm.

Motor structural integrity is a prerequisite for chemoreceptor self-assembly

Cryo-electron microscopy has successfully revealed the complete flagella structure in various bacterial species. This structure includes multiple components such as the inner-membrane MS ring, the cytoplasmic C ring, and the internal protein secretion system, all of which exhibit strong similarities34. This suggests that the core structure of the flagellar motor is highly conserved. The MS ring, composed of 26 FliF proteins, forms the foundation of the flagellar structure35. In addition to serving as a mounting platform for the C ring, the MS ring also acts as a protective shell for the internal protein secretion machinery36. FliG, the main component protein of the C ring, binds directly to the cytoplasmic surface of FliF in a 1:1 ratio37. Thus, the efficient assembly of the C ring requires preassembly of the MS ring. To observe the distribution of the chemosensory array following the disruption of the C ring and MS ring, we constructed ΔfliG and ΔfliF mutants.

We performed fluorescence observation on the ΔfliG mutant, which has a disrupted flagellar motor C ring. The results showed that the proportion of ΔfliG cells with obvious chemoreceptor clusters decreased significantly compared to the wild-type cells (59.8%, 140/234), although the chemosensory arrays remained at a single cell pole. We also examined the ΔfliF mutant, which has a disrupted flagellar motor MS ring. Similar to the ΔfliG mutant, the proportion of individuals with obvious chemoreceptor clusters decreased significantly (62.5%, 202/323). Subsequently, we constructed a ΔflhFΔfliF mutant and observed that the proportion of individuals with obvious chemoreceptor clusters further decreased (50.7%, 341/672). To ascertain whether it is motor integrity rather than functionality that influences the efficiency of chemosensory array assembly, we constructed a stator mutant (ΔmotAΔmotCD). In this mutant, the motor is completely stalled while the structure remains intact. We found that the mutant performed similarly to the wild-type strain in terms of chemosensory array assembly (84.3% of individuals with obvious clusters, 204/242). The fluorescence imaging of receptor clusters for multiple mutants in this section is shown in Figure. 3A. We utilized Western blotting to measure the expression levels of CheY, which were found to be similar across the different strains (Fig. 3B). This further substantiated that the observed phenomenon is based on the structural integrity of the motor rather than the protein expression level. Additionally, we quantified the proportion of cells with receptor clusters and the absolute fluorescence intensity of individual clusters (Fig. 3C-D). Overall, our findings suggest that the polar anchor protein FlhF and the structural integrity of the motor are crucial for the formation of chemosensory arrays. The former guides them to the appropriate site, while the latter influences the assembly efficiency of the chemoreceptor complex.

A. Subcellular localization of CheY-EYFP in various P. aeruginosa strains. The scale bar is 10 μm. B. Western blot analysis was performed to detect CheY expression in various P. aeruginosa strains. β-actin was used as the reference protein. C. Occurrence probability of obvious chemosensory arrays in P. aeruginosa wild-type and several mutant strains. The proportion of individuals with obvious chemosensory arrays decreased significantly in the motor-incomplete strains (ΔfliF and ΔfliG), and this value was further reduced after flhF knockout (ΔfliFΔflhF). The ΔfliF and ΔmotAΔmotCD strains have a similar chemosensory array occurrence probability as the wild-type strain. D. Absolute fluorescence brightness of the chemosensory array in P. aeruginosa wild-type and several mutant strains.

Colocalization of chemosensory array and flagellar motor avoids cross-pathway interference in P. aeruginosa signal transduction

The distribution patterns of receptors associated with various signal transduction pathways in P. aeruginosa vary considerably. For instance, the biofilm formation-related receptor WspA, unlike the chemotaxis receptors discussed here, is distributed throughout the cell. This distribution is thought to enhance its sensitivity to mechanical perturbations of the cell membrane38. Previous studies speculated that the highly consistent position of chemoreceptor arrays and flagellar motors enhances bacterial chemotaxis performance28. However, a spatial separation between the chemoreceptors and flagellar motors should not significantly impact the temporal comparison in bacterial chemotaxis. The diffusion time of the phosphorylated chemotaxis regulatory protein CheY-P across the longest distance in a bacterial cell body (along the cell’s long axis) is approximately 100 ms39, 40, while the timescale for the chemotaxis temporal comparison is on the order of seconds41. Consequently, the potential significance of the colocalization of chemoreceptors and flagellar motors remains unclear.

The signal transduction pathways in E. coli are relatively simple, and the chemotaxis response regulator CheY-P affects only the regulation of motor switching11. In contrast, the intracellular signal transduction pathways in P. aeruginosa are more complex. We thus hypothesized that the co-localization of chemoreceptors and flagellar motors in P. aeruginosa ensures locally distributed CheY-P molecules, thereby eliminating the need for a high level of intracellular CheY-P and avoiding potential side effects on other signaling pathways.

To test the potential effect of an increased intracellular CheY-P level, we constructed the CheY expression plasmid cheY-pJN105, transformed it into the wild-type P. aeruginosa strain, and induced it with varying concentrations of arabinose. We observed that the higher the inducer concentration, the more pronounced the cell aggregation became in the field of view (Fig. 4A). The transition from planktonic individuals to aggregated communities is similar to biofilm formation, which is known to be accompanied by a significant increase in intracellular c-di-GMP levels.

A. The evolution of cell aggregation as the intracellular CheY concentration increases by induction with higher concentrations of arabinose. The scale bar is 10 μm. B. Quantitative characterization of intracellular c-di-GMP levels at different CheY concentrations. From top to bottom, they correspond to the ΔcheY strain (N=198), wildtype strain (N=249) and CheY overexpression strain (N=228), respectively.

To investigate whether a higher level of intracellular CheY-P increased the intracellular c-di-GMP level, we sought to detect the c-di-GMP level. We introduced the plasmid pCdrA-gfp into P. aeruginosa, using it as a c-di-GMP biosensor with the fluorescence intensity proportional to the c-di-GMP level42. To ensure the coexistence of the plasmids, the intracellular CheY concentration was controlled by the plasmid cheY-pME6032. We found that the average single-cell fluorescence intensity of the CheY overexpression strain was 71.8% higher than that of the wild-type strain, while the c-di-GMP level of the cheY deletion strain was 58.6% lower than that of the wild-type (Fig. 4B).

Therefore, the colocalization of chemosensory array and flagellar motor facilitates the precise regulation of cell swimming direction within the low intracellular CheY-P concentration threshold, using locally distributed CheY-P molecules. This helps to avoid the occurrence of the aforementioned cross-pathway interference.

Summary and Discussion

P. aeruginosa harbors multiple signal transduction systems that regulate flagella-mediated swimming motility (Che pathway), pili-mediated interface twitching motility (Pil pathway), and biofilm formation (Wsp pathway)16. These systems allow it to thrive in various ecological niches within complex external environments. Here, employing a combination of chromosomal fluorescent protein fusion and flagellar labeling techniques in living cells, we directly observed that the chemosensory array of P. aeruginosa Che pathway and the flagellar motor share a high degree of consistency in their assembly sites. We found that the chemosensory array, comprising Che proteins, was consistently located at the cell pole where the flagellum was positioned. Based on these observations, we deduced the construction mode of the chemotactic network and flagellar motor during the entire cell growth cycle (Fig. 5A). As the cell body matures and elongates, a new flagellum grows at the opposing cell pole, accompanied by the assembly of a fresh chemosensory array. The cell then undergoes division from the middle, generating two daughter cells, each equipped with fully functional motility and chemotaxis capabilities. This ensures robust inheritance of these chemotaxis and motility-related macromolecular machines.

A. The construction mode of the chemotaxis network and flagellar motor throughout the complete cell growth cycle. B. The proximal growth mode of the P. aeruginosa flagellar motor and receptor clusters will effectively regulate the spatial range of CheY action, thus avoiding unintentional cross-pathway regulation.

Based on our understanding of the regulation mechanism of flagellar position in P. aeruginosa, we constructed a ΔflhF mutant and found that the position distribution of its chemosensory array is actively regulated by FlhF-mediated molecular mechanisms, rather than being passively restricted by membrane curvature factors as seen in E. coli. A similar active regulation phenomenon was previously reported in Vibrio cholerae, with the corresponding molecular regulation mechanism detailed extensively28. The flagella and chemosensory arrays of newborn V. cholerae cells are located at the old cell pole, with the polar anchor protein HubP distributed at both cell poles. As the cell grows, HubP recruits the ParA homologue ParC. Subsequently, ParC recruits ParP to assemble a new chemosensory array at the new cell pole. Upon cell division, HubP relocates to the middle of the cell, ensuring its robust presences at both poles of the daughter cells. Here, HubP recruits ParA homologues including FlhG, which influence flagellar motor assembly. However, previous experiments confirmed that the cell cycle-dependent polar localization of ParC and chemotactic proteins is independent of the flagellar position regulatory protein FlhF. This suggests that the localization of flagellar motors and chemosensory arrays in V. cholerae is controlled separately through different pathways. In contrast, flagellar motors were no longer robustly distributed at the cell pole after flhF was knocked out in P. aeruginosa, whereas the assembly site of chemosensory array and flagellar motor remained remarkably consistent. This suggests that FlhF regulates the distribution patterns of both and implies an interaction between them.

We further constructed ΔfliG and ΔfliF mutants, which disrupted the assembly of the flagellar motor. Surprisingly, we observed that the assembly of chemosensory array was impaired under these conditions, with a significant decrease in the proportion of individuals exhibiting obvious Che protein fluorescent bright spots. In contrast, the chemosensory array assembly in the stator mutant (ΔmotAΔmotCD) was similar to that of the wild-type strain. This suggests that the integrity rather than functionality of the flagellar motor in P. aeruginosa is critical for the assembly of its chemosensory array, and its assembly efficiency is greatly reduced in the absence of flagellar motors. To eliminate the potential confounding effects of FlhF and flagellar motors, we constructed a ΔflhFΔfliF mutant strain. Under this condition, the proportion of individuals with obvious Che protein fluorescent bright spots dropped to ∼50%, indicating that passive regulation based on membrane curvature may still exist in P. aeruginosa, albeit to a lesser degree.

Is the remarkable consistency in the spatial distribution of these two independent structural units an unintentional outcome or a hidden mystery? By introducing a c-di-GMP monitoring system at the single-cell level, we discovered that overexpressing CheY substantially increases the intracellular c-di-GMP concentration. Previous studies generally believed that CheY-P, as a chemotactic regulatory protein, only influences flagellar motor switching. Here, we found that CheY-P, as a signaling molecule, can accomplish threshold-limited cross-pathway regulation. At low CheY-P concentrations, it is conservatively involved in motor switching regulation, as confirmed by previous studies. At high CheY-P concentrations, cell motility is suppressed, and c-di-GMP is triggered to express, though its specific mechanism needs further exploration. The intracellular environment is highly crowded43, and the proximal growth pattern shown in Fig. 5B has multiple physiological significances. On one hand, chemotaxis-related protein can precisely regulate cell motility with minimal synthesis costs. On the other hand, the chemotactic network and the flagellar motor are spatially integrated within the cell, ensuring their robust existence within the cell body and preventing mutual interference with collateral pathways.

Many bacteria possess multiple signal transduction pathways. The orderly operation of each pathway within the micrometer-scale space presents a fascinating scientific problem. Here, we utilized the well-known chemotaxis network as a starting point to shed light on this problem, providing inspiration for future in-depth understanding of related issues.

Materials and methods

Strains and cell culture

The strains and plasmids used in this study are listed in Table 1. The Escherichia coli TOP10 strain was used for standard genetic manipulations. A single-colony isolate was grown in 3 ml of LB broth (1% Bacto tryptone, 0.5% yeast extract and 1% NaCl) overnight to saturation on a rotary shaker (250 rpm) at 37°C. An aliquot was diluted 1:100 into 10 ml of LB broth and grew to exponential phase. Appropriate antibiotics were added if necessary to prevent plasmid loss: for E. coli, 15 μg/ml Gentamicin and 25 μg/ml Tetracycline; for P. aeruginosa, 30 μg/ml Gentamicin and 50 μg/ml Tetracycline. To induce protein expression, Isopropyl β-D-1-thiogalactopyranoside (IPTG, 1.0 mM) was added to strains with pME6032-derivative vectors, and arabinose (0.005%, 0.01%, 0.05%, 0.1%) was added to strains with pJN105-derivative vectors. 2 ml of cells were harvested by centrifugation at 2000×g for 2 min, washed twice in an equal volume of motility buffer (MB) [50 mM potassium phosphate, 15 μM EDTA, 0.15 M NaCl, 5 mM Mg2+ and 10 mM lactic acid (pH 7.0)]44, and resuspended in 5 ml MB for subsequent fluorescence observation.

Construction of in-frame deletion mutants

We used polymerase chain reaction (PCR) to generate ∼1000 bp DNA fragments with upstream (Up) and downstream (Dn) sequences flanking the target gene (including flhF, fliF, motA and fliG) to be deleted. The Up and Dn DNA sequences were fused with the linearized pex18gm vector using Gibson assembly45. The resulting vectors were transformed into P. aeruginosa by electroporation, and the desired knockout mutants were obtained by double selections on gentamicin plates (LB plates with 30 µg/ml gentamicin) and sucrose plates (NaCl-free LB plates with 15% sucrose) at 37°C46.

Generation of chromosomal fusions of yfp to P. aeruginosa gene

To generate eyfp fusion at the N-terminus of cheY, we first used PCR to generate upstream and downstream DNA sequences flanking the cheY gene. A subsequent PCR reaction aimed to yield the cheY-eyfp fusion with a 3× glycine linker (GGCGGAGGA), with pVS88 serving as the source of the enhanced yellow fluorescent (eyfp). The three DNA sequences, along with the linearized vector, were fused to cheY-eyfp-pex18gm using Gibson assembly. Finally, gene replacement was used to transfer the fusion constructs into the P. aeruginosa chromosome by homologous recombination.

Flagellar staining and fluorescence imaging

Flagellar filaments were labeled by following the protocol described previously47. Cells (1 ml of exponential-phase culture) were harvested by centrifugation at 2000×g for 10 min and washed twice in 1 ml MB. The final pellet was adjusted to a volume of ∼100 µl that concentrated the bacterial 10-fold. Alexa Fluor 568 maleimide (Invitrogen-Molecular Probes) was added to a final concentration of 20 μg/ml, and labeling was allowed to proceed for 30 min at room temperature with gyration at 80 rpm. Unused dye was then removed by washing the cells with MB three times, and the final pellet was resuspended in MB.

For fluorescence imaging, 50 μl of cells were added to the chamber (constructed with two double-sticky tapes, a glass slide, and a poly-L-Lysine coated coverslip), incubated for 3 min, and rinsed with 100 μl MB. The boundary of the chamber was then sealed with Apiezon vacuum grease. The chamber was placed on a Nikon Ti-E inverted fluorescence microscope with a 100× oil-immersion objective and a sCMOS camera (Primer95B, Photometrics). The flagellar filament and chemosensory array of cells were observed separately with corresponding filter-set, using a 200-ms exposure time. To quantify the fluorescence intensity of the receptor cluster, we took the intensity maximum point as the center and extract a 3×3 pixel matrix around it. The mean value of the elements in this matrix represents the final fluorescence intensity.

Western Blotting

The total protein of P. aeruginosa strain cells in each treatment group was extracted and its concentration was measured with Bicinchoninic acid (BCA Protein Quantification Kit, Yeasen Biotechnology Co., Ltd) when the cells grew to exponential phase with same reduced turbidity (OD600 between 0.9 and 1.0). The exact amount of protein was subjected to SDS-PAGE electrophoresis, then transferred to a nitrocellulose membrane, which was blocked with 50 mg/mL skim milk in TBST buffer (20 mM Tris, 150 mM NaCl [pH7.4], 0.1%Tween 20) for 1h at room temperature. Since the N-terminal sequence of EYFP is almost identical to that of GFP, a rabbit anti-GFP antibody (1:2000, Yeasen Biotechnology Co., Ltd) and a Mouse anti-β-actin antibody (1:2000, Invitrogen) were added and incubated at 4°C overnight. The nitrocellulose membrane was gently washed with TBST for 20 min, three times. Subsequently, HRP labeled goat anti-rabbit IgG (1:4000, Abcam) and rabbit anti-mouse IgG (1:4000, Abcam) were added and incubated at room temperature for 1 h. The nitrocellulose membrane was again gently washed with TBST for 20 min, three times. Then, the membrane was developed using the ECL chemiluminescence detection system (Beyotime P0018FS) and visualized with the ChemiDoc XRS+ system (Bio-Rad). Three independent experiments were conducted for reproducibility.

Monitoring c-di-GMP signal at the single-cell level

For measurements of c-di-GMP signaling, the validated reporter plasmid pCdrA-gfp, which produces green fluorescent protein (GFP) in response to an increase in c-di-GMP, was introduced into the experimental system. A 1000× diluted culture was inoculated into the apparatus described in the previous section and allowed to adhere to the surface for ∼10 min, with the slight difference that the coverslips were not coated with poly-L-Lysine to avoid interference from background fluorescence. Bacteria were imaged using a Nikon Ti-E inverted microscope, which was equipped with an EMCCD camera (DU897, iXon3, Andor Technology). We used a 100× oil-immersion objective. Bacteria were illuminated with a 488-nm laser (Sapphire 488−200 mW, Coherent), using a standard GFP filter sets in the fluorescence imaging system with an exposure time of 500 ms. To ensure the consistency of culture and observation conditions, the data of wild-type and cheY-pME6032 carried strains were collected at the same time. More than 200 independent individuals of each strain were used for data analysis.

Single-cell c-di-GMP concentration was characterized by calculating fluorescence intensity. The average fluorescence intensity of a single cell was calculated by dividing the total fluorescence intensity of a single cell by the cell volume. Bacterial were simplified to a hollow cylindrical trunk with hollow hemispherical caps at both ends. The cell volume can be calculated by taking the length of the short axis of the cell body as the diameter of the sphere and the length of the long axis as the sum of the height of the cylinder and the diameter of the sphere. The mean fluorescence intensity of the cell was obtained by subtracting the average background fluorescence intensity from the total fluorescence intensity of the cell, and then dividing by the cell volume.

eLife Assessment

This valuable study by Wu et al presents data on bacterial cell organization, with the goal of demonstrating that the two structures that account for bacterial motility - the chemotaxis complex and the flagella - colocalize to the same pole in Pseudomonas aeruginosa cells and expose the regulation underlying their spatial organization and functioning. The subject is of importance and the manuscript is well written. However, the work is incomplete and the conclusions are too strong for the presented data. This manuscript will be of interest to cell biologists, mainly those studying bacteria, but not only, if the evidence is substantiated, the assumptions clarified, and the novelty is made clear during the revision process.

Reviewer #1 (Public Review):

Summary:

The study by Wu et al presents interesting data on bacterial cell organization, a field that is progressing now, mainly due to the advances in microscopy. Based mainly on fluorescence microscopy images, the authors aim to demonstrate that the two structures that account for bacterial motility, the chemotaxis complex and the flagella, colocalize to the same pole in Pseudomonas aeruginosa cells and to expose the regulation underlying their spatial organization and functioning.

Strengths:

The subject is of importance.

Weaknesses:

The conclusions are too strong for the presented data. The lack of statistical analysis makes this paper incomplete. The novelty of the findings is not clear.

Major issues:

(1) The novelty is in question since in the Abstract the authors highlight their main finding, which is that both the chemotaxis complex and the flagella localize to the same pole, as surprising. However, in the Introduction they state that "pathway-related receptors that mediate chemotaxis, as well as the flagellum are localized at the same cell pole17,18". I am not a pseudomonas researcher and from my short glance at these references, I could not tell whether they report colocalization of the two structures to the same pole. However, I trust the authors that they know the literature on the localization of the chemotaxis complex and flagella in their organism. See also major issue number 5 on the novelty regarding the involvement of c-di-GMP.

(2) Statistics for the microscopy images, on which most conclusions in this manuscript are based, are completely missing. Given that most micrographs present one or very few cells, together with the fact that almost all conclusions depend on whether certain macromolecules are at one or two poles and whether different complexes are in the same pole, proper statistics, based on hundreds of cells in several fields, are absolutely required. Without this information, the results are anecdotal and do not support the conclusions. Due to the importance of statistics for this manuscript, strict statistical tests should be used and reported. Moreover, representative large fields with many cells should be added as supportive information.

The problem is more pronounced when the authors make strong statements, as in lines 157-158: "The results revealed that the chemoreceptor arrays no longer grow robustly at the cell pole (Figure 2A)". Looking at the seven cells shown in Figure 2A, five of them show polar localization of the chemoreceptors. The question is then: what is the percentage of cells that show precise polar, near-polar, or mid cell localization (the three patterns shown here) in the mutant and in the wild type? Since I know that these three patterns can also be observed in WT cells, what counts is the difference, and whether it is statistically significant.

Even for the graphs shown in Figures 3C and 3D, where the proportion of cells with obvious chemoreceptor arrays and absolute fluorescence brightness of the chemosensory array are shown, respectively, the questions that arise are: for how many individual cells these values hold and what is the significance of the difference between each two strains?

(3) The authors conclude that "Motor structural integrity is a prerequisite for chemoreceptor self-assembly" based on the reduction in cells with chemoreceptor clusters in mutants deleted for flagellar genes, despite the proper polar localization of the chemotaxis protein CheY. They show that the level of CheY in the WT and the mutant strains is similar, based on Western blot, which in my opinion is over-exposed. "To ascertain whether it is motor integrity rather than functionality that influences the efficiency of chemosensory array assembly", they constructed a mutant deleted for the flagella stator and found that the motor is stalled while CheY behaves like in WT cells. The authors further "quantified the proportion of cells with receptor clusters and the absolute fluorescence intensity of individual clusters (Figures 3C-D)". While Figure 3DC suggests that, indeed, the flagella mutants show fewer cells with a chemotaxis complex, Figure 3D suggests that the differences in fluorescence intensity are not statistically significant.

Since it is obvious that the regulation of both structures' production and localization is codependent, I think that it takes more than a Western blot to make such a decision.

(4) I wonder why the authors chose to label CheY, which is the only component of the chemotaxis complex that shuttles back and forth to the base of the flagella. In any case, I think that they should strengthen their results by repeating some key experiments with labeled CheW or CheA.

(5) The last section of the results is very problematic, regarding the rationale, the conclusions, and the novelty. As far as the rationale is concerned, I do not understand why the authors assume that "a spatial separation between the chemoreceptors and flagellar motors should not significantly impact the temporal comparison in bacterial chemotaxis". Is there any proof for that? More surprising for me was to read that "The signal transduction pathways in E. coli are relatively simple, and the chemotaxis response regulator CheY-P affects only the regulation of motor switching". There are degrees of complexity among signal transduction pathways in E. coli, but the chemotaxis seems to be ranked at the top. CheY is part of the adaptation. Perfect adaptation, as many other issues related to the chemotaxis pathway, which include the wide dynamic range, the robustness, the sensitivity, and the signal amplification (gain), are still largely unexplained. Hence, such assumptions are not justified.

More perplexing is the novelty of the authors' documentation of the effect of the chemotaxis proteins on the c-di-GMP level. In 2013, Kulasekara et al. published a paper in eLife entitled "c-di-GMP heterogeneity is generated by the chemotaxis machinery to regulate flagellar motility". In the same year, Kulasekara published a paper entitled "Insight into a Mechanism Generating Cyclic di-GMP Heterogeneity in Pseudomonas aeruginosa". The authors did not cite these works and I wonder why.

(6) Throughout the manuscript, the authors refer to foci of fluorescent CheY as "chemoreceptor arrays". If anything, these foci signify the chemotaxis complex, not the membrane-traversing chemoreceptors.

Conclusions:

The manuscript addresses an interesting subject and contains interesting, but incomplete, data.

Reviewer #2 (Public Review):

Summary:

Here, the authors studied the molecular mechanisms by which the chemoreceptor cluster and flagella motor of Pseudomonas aeruginosa (PA) are spatially organized in the cell. They argue that FlhF is involved in localizing the receptors-motor to the cell pole, and even without FlhF, the two are colocalized. FlhF is known to cause the motor to localize to the pole in a different bacterial species, Vibrio cholera, but it is not involved in receptor localization in that bacterium. Finally, the authors argue that the functional reason for this colocalization is to insulate chemotactic signaling from other signaling pathways, such as cyclic-di-GMP signaling.

Strengths:

The experiments and data look to be high-quality.

Weaknesses:

However, the interpretations and conclusions drawn from the experimental observations are not fully justified in my opinion.

I see two main issues with the evidence provided for the authors' claims.

(1) Assumptions about receptor localization:

The authors rely on YFP-tagged CheY to identify the location of the receptor cluster, but CheY is a diffusible cytoplasmic protein. In E. coli, CheY has been shown to localize at the receptor cluster, but the evidence for this in PA is less strong. The authors refer to a paper by Guvener et al 2006, which showed that CheY localizes to a cell pole, and CheA (a receptor cluster protein) also localizes to a pole, but my understanding is that colocalization of CheY and CheA was not shown. My concern is that CheY could instead localize to the motor in PA, say by binding FliM. This "null model" would explain the authors' observations, without colocalization of the receptors and motor.

Verifying that CheY and CheA are colocalized in PA would be a very helpful experiment to address this weakness.

(2) Argument for the functional importance of receptor-motor colocalization at the pole:

The authors argue that colocalization of the receptors and motors at the pole is important because it could keep phosphorylated CheY, CheY-p, restricted to a small region of the cell, preventing crosstalk with other signaling pathways. Their evidence for this is that overexpressing CheY leads to higher intracellular cdG levels and cell aggregation.

Say that the receptors and motors are colocalized at the pole. In E. coli, CheY-p rapidly diffuses through the cell. What would prevent this from occurring in PA, even with colocalization?

Elevating CheY concentration may increase the concentration of CheY-p in the cell, but might also stress the cells in other unexpected ways. It is not so clear from this experiment that elevated CheY-p throughout the cell is the reason that they aggregate, or that this outcome is avoided by colocalizing the receptors and motor at the same pole.

If localization of the receptor array and motor at one pole were important for keeping CheY-p levels low at the opposite pole, then we should expect cells in which the receptors and motor are not at the pole to have higher CheY-p at the opposite pole. According to the authors' argument, it seems like this should cause elevated cdG levels and aggregation in the delta flhF mutants with wild-type levels of CheY. But it does not look like this happened.

Instead of varying CheY expression, the authors could test their hypothesis that receptor-motor colocalization at the pole is important for preventing crosstalk by measuring cdG levels in the flhF mutant, in which the motor (and maybe the receptor cluster) are no longer localized in the cell pole.

Reviewer #3 (Public Review):

Summary:

The authors investigated the assembly and polar localization of the chemosensory cluster in P. aeruginosa. They discovered that a certain protein (FlhF) is required for the polar localization of the chemosensory cluster while a fully-assembled motor is necessary for the assembly of the cluster. They found that flagella and chemosensory clusters always co-localize in the cell; either at the cell pole in wild-type cells or randomly-located in the cell in FlhF mutant cells. They hypothesize that this co-localization is required to keep the level of another protein (CheY-P), which controls motor switching, at low levels as the presence of high levels of this protein (if the flagella and chemosensory clusters were not co-localized) is associated with high-levels of c-di-GMP and cell aggregations.

Strengths:

The manuscript is clearly written and straightforward. The authors applied multiple techniques to study the bacterial motility system including fluorescence light microscopy and gene editing. In general, the work enhances our understanding of the subtlety of interaction between the chemosensory cluster and the flagellar motor to regulate cell motility.

Weaknesses:

The major weakness in this paper is that the authors never discussed how the flagellar gene expression is controlled in P. aeruginosa. For example, in E. coli there is a transcriptional hierarchy for the flagellar genes (early, middle, and late genes, see Chilcott and Hughes, 2000). Similarly, Campylobacter and Helicobacter have a different regulatory cascade for their flagellar genes (See Lertsethtakarn, Ottemann, and Hendrixson, 2011). How does the expression of flagellar genes in P. aeruginosa compare to other species? How many classes are there for these genes? Is there a hierarchy in their expression and how does this affect the results of the FliF and FliG mutants? In other words, if FliF and FliG are in class I (as in E. coli) then their absence might affect the expression of other later flagellar genes in subsequent classes (i.e., chemosensory genes). Also, in both FliF and FliG mutants no assembly intermediates of the flagellar motor are present in the cell as FliG is required for the assembly of FliF (see Hiroyuki Terashima et al. 2020, Kaplan et al. 2019, Kaplan et al. 2022). It could be argued that when the motor is not assembled then this will affect the expression of the other genes (e.g., those of the chemosensory cluster) which might play a role in the decreased level of chemosensory clusters the authors find in these mutants.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The study by Wu et al presents interesting data on bacterial cell organization, a field that is progressing now, mainly due to the advances in microscopy. Based mainly on fluorescence microscopy images, the authors aim to demonstrate that the two structures that account for bacterial motility, the chemotaxis complex and the flagella, colocalize to the same pole in Pseudomonas aeruginosa cells and to expose the regulation underlying their spatial organization and functioning.

Strengths:

The subject is of importance.

Weaknesses:

The conclusions are too strong for the presented data. The lack of statistical analysis makes this paper incomplete. The novelty of the findings is not clear.

We have strengthened the data analysis by including appropriate statistical tests to support our conclusions more convincingly. Additionally, we have refined the description of the research background to better emphasize the novelty and significance of our findings. Please see the detailed responses below for further information.

Major issues:

(1) The novelty is in question since in the Abstract the authors highlight their main finding, which is that both the chemotaxis complex and the flagella localize to the same pole, as surprising. However, in the Introduction they state that "pathway-related receptors that mediate chemotaxis, as well as the flagellum are localized at the same cell pole17,18". I am not a pseudomonas researcher and from my short glance at these references, I could not tell whether they report colocalization of the two structures to the same pole. However, I trust the authors that they know the literature on the localization of the chemotaxis complex and flagella in their organism. See also major issue number 5 on the novelty regarding the involvement of c-di-GMP.

We thank the reviewer for this valuable comment and appreciate the opportunity to clarify our statements.

Kazunobu et al. (ref. 18) used scanning electron microscopy to preliminarily characterize the flagellation pattern of Pseudomonas aeruginosa during cell division, showing that existing flagella are located at the old pole. Zehra et al. (ref. 17), through fluorescence microscopy, observed that CheA and CheY proteins in dividing cells are typically also present at the old pole. Based on these observations, we inferred in the Introduction that the chemotaxis complex and flagellum may localize to the same cell pole.

However, this inference is indirect and lacks direct live-cell evidence of colocalization, leaving its validity to be confirmed. This uncertainty was indeed the starting point and motivation for our study.

In our work, we simultaneously visualized flagellar filaments and core chemoreceptor proteins at the single-cell level in P. aeruginosa. We characterized the assembly and spatial coordination of the chemotaxis network and flagellar motor throughout the cell cycle, providing direct evidence of their colocalization and coordinated assembly. This represents a significant advance beyond prior indirect observations and supports the novelty of our study.

Accordingly, we have revised the relevant statements in lines 71-75 of the manuscript to better reflect the current state of the literature and emphasize the novelty of our direct observations.

(2) Statistics for the microscopy images, on which most conclusions in this manuscript are based, are completely missing. Given that most micrographs present one or very few cells, together with the fact that almost all conclusions depend on whether certain macromolecules are at one or two poles and whether different complexes are in the same pole, proper statistics, based on hundreds of cells in several fields, are absolutely required. Without this information, the results are anecdotal and do not support the conclusions. Due to the importance of statistics for this manuscript, strict statistical tests should be used and reported. Moreover, representative large fields with many cells should be added as supportive information.

We thank the reviewer for this important comment, which significantly improves the rigor and persuasiveness of our manuscript.

For the colocalization analyses presented in Fig. 1D and Fig. 2B, we quantified 145 and 101 cells with fluorescently labeled flagella, respectively, and observed consistent colocalization of the chemoreceptor complexes and flagella in all examined cells (now added in the figure legends). Regarding the distribution patterns of chemoreceptors shown in Fig. 3A, we have now included comprehensive statistical analyses for both wild-type and mutant strains. For each strain, more than 300 cells were analyzed across at least three independent microscopic fields, providing robust statistical power (detailed data are presented in Fig. 3C).

To further strengthen the evidence, statistical tests were applied to confirm the significance and reproducibility of our findings (Fig. 3C). In addition, representative large-field fluorescence images containing numerous cells have been added to the supplementary materials (Fig. S1 and Fig. S3).

The problem is more pronounced when the authors make strong statements, as in lines 157-158: "The results revealed that the chemoreceptor arrays no longer grow robustly at the cell pole (Figure 2A)". Looking at the seven cells shown in Figure 2A, five of them show polar localization of the chemoreceptors. The question is then: what is the percentage of cells that show precise polar, near-polar, or mid cell localization (the three patterns shown here) in the mutant and in the wild type? Since I know that these three patterns can also be observed in WT cells, what counts is the difference, and whether it is statistically significant.

We thank the reviewer for raising this important point. Following the reviewer's suggestion, we have now analyzed and categorized the distribution of the chemotaxis complex in both wild-type and flhF mutant strains into three patterns: precise-polar, near-polar, and mid-cell localization. For each strain, more than 200 cells across three independent fields of view were quantified.

Our statistical analysis shows that in the wild-type strain, approximately 98% of cells exhibit precise polar localization of the chemotaxis complex. In contrast, the ΔflhF mutant displays a clear shift in distribution, with about 5% of cells showing mid-cell localization and 9.5% showing near-polar localization. These differences demonstrate a significant alteration in the spatial pattern upon flhF deletion.

We have revised the relevant text in lines 166-170 accordingly and included the detailed statistical data in the newly added Fig. S4.

Even for the graphs shown in Figures 3C and 3D, where the proportion of cells with obvious chemoreceptor arrays and absolute fluorescence brightness of the chemosensory array are shown, respectively, the questions that arise are: for how many individual cells these values hold and what is the significance of the difference between each two strains?

The number of cells analyzed for each strain is indicated in the original manuscript: 372 wild-type cells (line 123), 221 ΔflhF cells (line 172), 234 ΔfliG cells (line 197), 323 ΔfliF cells (line 200), 672 ΔflhFΔfliF cells (line 202), and 242 ΔmotAΔmotCD cells (line 207). For each strain, data were collected from three independent fields of view. We have now also provided the number of cells in Fig. 3 legend.

We have now performed statistical comparisons using t-tests between strains. Notably, the measured values in Fig. 3C exhibit a clear, monotonic decrease with successive gene knockouts, supporting the robustness of the observed trend.

Regarding the absolute fluorescence intensity shown in the original Fig. 3D, the mutants did not display consistent directional changes compared to the wild type. Reliable comparison of absolute fluorescence intensity requires consistent fluorescent protein maturation levels across strains. Given the likely variability in maturation levels between strains, we concluded that this data may not accurately reflect true differences in protein concentrations. Therefore, we have removed the fluorescence intensity graph from the revised manuscript to avoid potential misinterpretation.

(3) The authors conclude that "Motor structural integrity is a prerequisite for chemoreceptor self-assembly" based on the reduction in cells with chemoreceptor clusters in mutants deleted for flagellar genes, despite the proper polar localization of the chemotaxis protein CheY. They show that the level of CheY in the WT and the mutant strains is similar, based on Western blot, which in my opinion is over-exposed. "To ascertain whether it is motor integrity rather than functionality that influences the efficiency of chemosensory array assembly", they constructed a mutant deleted for the flagella stator and found that the motor is stalled while CheY behaves like in WT cells. The authors further "quantified the proportion of cells with receptor clusters and the absolute fluorescence intensity of individual clusters (Figures 3C-D)". While Figure 3DC suggests that, indeed, the flagella mutants show fewer cells with a chemotaxis complex, Figure 3D suggests that the differences in fluorescence intensity are not statistically significant. Since it is obvious that the regulation of both structures' production and localization is codependent, I think that it takes more than a Western blot to make such a decision.

We thank the reviewer for the suggestions. To further clarify that the assembly of flagellar motors and chemoreceptor clusters occurs in an orderly manner rather than being merely codependent, we performed additional experiments. Specifically, we constructed a ΔcheA mutant strain, in which chemoreceptor clusters fail to assemble. Using in vivo fluorescent labeling of flagellar filaments, we observed that the proportion of cells with flagellar filaments in the ΔcheA strain was comparable to that of the wild type (Fig. S5).

In contrast, mutants lacking complete motor structures, such as ΔfliF and ΔfliG, showed a significant reduction in the proportion of cells with obvious receptor clusters (Fig. 3C). Based on these results, we conclude that the structural integrity of the flagellar motor is, to a certain extent, a prerequisite for the self-assembly of chemoreceptor clusters.

Accordingly, we have revised the relevant statement in lines 213-217 of the manuscript to reflect this clarification.

(4) I wonder why the authors chose to label CheY, which is the only component of the chemotaxis complex that shuttles back and forth to the base of the flagella. In any case, I think that they should strengthen their results by repeating some key experiments with labeled CheW or CheA.

We thank the reviewer for this valuable suggestion. In our study, we initially focused on the positional relationship between chemoreceptor clusters and flagella, then investigated factors influencing cluster distribution and assembly efficiency. The physiological significance of motor and cluster co-localization was ultimately proposed with CheY as the starting point.

Previous work by Harwood's group demonstrated that both CheY-YFP and CheA-GFP localize to the old poles of dividing Pseudomonas aeruginosa cells. Since our physiological hypothesis centers on CheY, we chose to label CheY-EYFP in our experiments.

To further strengthen our conclusions, we constructed a plasmid expressing CheA-CFP and introduced it into the cheY-eyfp strain via electroporation. Fluorescence imaging revealed a high degree of spatial overlap between CheA-CFP and CheY-EYFP (Fig. S2), confirming that CheY-EYFP accurately marks the location of the chemoreceptor complex.

We have revised the manuscript accordingly (lines 119-123) and added these data as Fig. S2.

(5) The last section of the results is very problematic, regarding the rationale, the conclusions, and the novelty. As far as the rationale is concerned, I do not understand why the authors assume that "a spatial separation between the chemoreceptors and flagellar motors should not significantly impact the temporal comparison in bacterial chemotaxis". Is there any proof for that?

We apologize for the lack of clarity in our original explanation. The rationale behind the statement was initially supported by comparing the timescales of CheY-P diffusion and temporal comparison in chemotaxis. Specifically, the diffusion time for CheY-P to traverse the entire length of a bacterial cell is approximately 100 ms (refs 39&40), whereas the timescale for bacterial chemotaxis temporal comparison is on the order of seconds (ref 41).

To clarify and strengthen this argument, we have expanded the discussion as follows:

The diffusion coefficient of CheY in bacterial cells is about 10 µm2/s, which corresponds to an estimated end-to-end diffusion time on the order of 100 ms (refs 40&41). If the chemotaxis complexes were randomly distributed rather than localized, diffusion times would be even shorter. In contrast, the timescale for the chemotaxis temporal comparison is on the order of seconds (ref. 42). Additionally, a study by Fukuoka and colleagues reported that intracellular chemotaxis signal transduction requires approximately 240 ms beyond CheY or CheY-P diffusion time (ref. 41). Moreover, the intervals of counterclockwise (CCW) and clockwise (CW) rotation of the P. aeruginosa flagellar motor under normal conditions are 1-2 seconds, as determined by tethered cell or bead assays (refs. 30&43).

Taken together, these indicate that for P. aeruginosa, which moves via a run-reverse mode, the potential 100 ms reduction in response time due to co-localization of the chemotaxis complex and motor has a limited effect on overall chemotaxis timing.

We have revised the corresponding text accordingly (lines 238-245) to better explain this rationale.

More surprising for me was to read that "The signal transduction pathways in E. coli are relatively simple, and the chemotaxis response regulator CheY-P affects only the regulation of motor switching". There are degrees of complexity among signal transduction pathways in E. coli, but the chemotaxis seems to be ranked at the top. CheY is part of the adaptation. Perfect adaptation, as many other issues related to the chemotaxis pathway, which include the wide dynamic range, the robustness, the sensitivity, and the signal amplification (gain), are still largely unexplained. Hence, such assumptions are not justified.

We apologize for the confusion and imprecision in our original statements. Our intention was to convey that the chemotaxis pathway in E. coli is relatively simple compared to the more complex chemosensory systems in P. aeruginosa. We did not mean to generalize this simplicity to all signal transduction pathways in E. coli.

We acknowledge that E. coli chemotaxis is a highly sophisticated system, involving processes such as perfect adaptation, wide dynamic range, robustness, sensitivity, and signal amplification, many aspects of which remain incompletely understood. CheY indeed plays a crucial role in adaptation and motor switching regulation.

Accordingly, we have revised the original text (lines 249-255) to avoid any misunderstanding.

More perplexing is the novelty of the authors' documentation of the effect of the chemotaxis proteins on the c-di-GMP level. In 2013, Kulasekara et al. published a paper in eLife entitled "c-di-GMP heterogeneity is generated by the chemotaxis machinery to regulate flagellar motility". In the same year, Kulasekara published a paper entitled "Insight into a Mechanism Generating Cyclic di-GMP Heterogeneity in Pseudomonas aeruginosa". The authors did not cite these works and I wonder why.

We apologize for having been unaware of these important references and thank the reviewer for bringing them to our attention. We have now cited the eLife paper and the PhD thesis titled "Insight into a Mechanism Generating Cyclic di-GMP Heterogeneity in Pseudomonas aeruginosa" by Kulasekara et al.

Regarding novelty, there are key differences between our findings and those reported by Kulasekara et al. While they proposed that CheA influences c-di-GMP heterogeneity through interaction with a specific phosphodiesterase (PDE), our results demonstrate that overexpression of CheY leads to an increase in intracellular c-di-GMP levels.

We have revised the original text accordingly (lines 358-362) to clarify these distinctions.

(6) Throughout the manuscript, the authors refer to foci of fluorescent CheY as "chemoreceptor arrays". If anything, these foci signify the chemotaxis complex, not the membrane-traversing chemoreceptors.

We thank the reviewer for this clarification. We have revised the manuscript accordingly to refer to the fluorescent CheY foci as representing the chemotaxis complex rather than the chemoreceptor arrays.

Conclusions:

The manuscript addresses an interesting subject and contains interesting, but incomplete, data.

Reviewer #2 (Public Review):

Summary:

Here, the authors studied the molecular mechanisms by which the chemoreceptor cluster and flagella motor of Pseudomonas aeruginosa (PA) are spatially organized in the cell. They argue that FlhF is involved in localizing the receptors-motor to the cell pole, and even without FlhF, the two are colocalized. FlhF is known to cause the motor to localize to the pole in a different bacterial species, Vibrio cholera, but it is not involved in receptor localization in that bacterium. Finally, the authors argue that the functional reason for this colocalization is to insulate chemotactic signaling from other signaling pathways, such as cyclic-di-GMP signaling.

Strengths:

The experiments and data look to be high-quality.

Weaknesses:

However, the interpretations and conclusions drawn from the experimental observations are not fully justified in my opinion.

I see two main issues with the evidence provided for the authors' claims.

(1) Assumptions about receptor localization:

The authors rely on YFP-tagged CheY to identify the location of the receptor cluster, but CheY is a diffusible cytoplasmic protein. In E. coli, CheY has been shown to localize at the receptor cluster, but the evidence for this in PA is less strong. The authors refer to a paper by Guvener et al 2006, which showed that CheY localizes to a cell pole, and CheA (a receptor cluster protein) also localizes to a pole, but my understanding is that colocalization of CheY and CheA was not shown. My concern is that CheY could instead localize to the motor in PA, say by binding FliM. This "null model" would explain the authors' observations, without colocalization of the receptors and motor. Verifying that CheY and CheA are colocalized in PA would be a very helpful experiment to address this weakness.

We thank the reviewer for this valuable suggestion. We agree that verifying the colocalization of CheY and CheA would strengthen our conclusions. To address this, we constructed a plasmid expressing CheA-CFP and introduced it into the CheY-EYFP strain by electroporation. Fluorescence imaging revealed a high degree of spatial overlap between CheA-CFP and CheY-EYFP signals, indicating that CheY-EYFP indeed marks the location of the chemoreceptor complex rather than the flagellar motor.

We have revised the manuscript accordingly (lines 118-123) and included these results in the new Fig. S2.

(2) Argument for the functional importance of receptor-motor colocalization at the pole:

The authors argue that colocalization of the receptors and motors at the pole is important because it could keep phosphorylated CheY, CheY-p, restricted to a small region of the cell, preventing crosstalk with other signaling pathways. Their evidence for this is that overexpressing CheY leads to higher intracellular cdG levels and cell aggregation. Say that the receptors and motors are colocalized at the pole. In E. coli, CheY-p rapidly diffuses through the cell. What would prevent this from occurring in PA, even with colocalization?

We appreciate the reviewer's insightful question. The colocalization of both the signaling source (the kinase) and sink (the phosphatase) at the chemoreceptor complex at the cell pole results in a rapid decay of CheY-P concentration within approximately 0.2 µm from the cell pole, leading to a nearly uniform distribution elsewhere in the cell, as demonstrated by Vaknin and Berg (ref. 46). This spatial arrangement effectively confines high CheY-P levels to the pole region. When the motor is also localized at the cell pole, this reduces the need for elevated CheY-P concentrations throughout the cytoplasm, thereby minimizing potential crosstalk with other signaling pathways.

We have revised the manuscript accordingly (lines 280-286) to clarify this point.

Elevating CheY concentration may increase the concentration of CheY-p in the cell, but might also stress the cells in other unexpected ways. It is not so clear from this experiment that elevated CheY-p throughout the cell is the reason that they aggregate, or that this outcome is avoided by colocalizing the receptors and motor at the same pole. If localization of the receptor array and motor at one pole were important for keeping CheY-p levels low at the opposite pole, then we should expect cells in which the receptors and motor are not at the pole to have higher CheY-p at the opposite pole. According to the authors' argument, it seems like this should cause elevated cdG levels and aggregation in the delta flhF mutants with wild-type levels of CheY. But it does not look like this happened. Instead of varying CheY expression, the authors could test their hypothesis that receptor-motor colocalization at the pole is important for preventing crosstalk by measuring cdG levels in the flhF mutant, in which the motor (and maybe the receptor cluster) are no longer localized in the cell pole.

We thank the reviewer for raising the important point regarding potential cellular stress caused by elevated CheY concentrations, as well as for the suggestion to test the hypothesis using ΔflhF mutants.

First, as noted above, CheY-P concentration rapidly decreases away from the receptor complex. While deletion of flhF alters the position of the receptor complex, thereby shifting the region of high CheY-P concentration, it does not increase CheY-P levels elsewhere in the cell. Importantly, in the ΔflhF strain, the receptor complex and the motor still colocalize, so this mutant may not effectively test the role of receptor-motor colocalization in preventing crosstalk as suggested.

Regarding the possibility that elevated CheY levels stress the cells independently of CheY-P signaling, prior work in <i.E. coli by Cluzel et al. (ref. 11) showed that overexpressing CheY several-fold did not cause phenotypic changes, indicating that simple CheY overexpression alone may not be generally stressful. Furthermore, our data indicate that the increase in c-di-GMP levels and subsequent cell aggregation upon CheY overexpression is not an all-or-none switch but occurs progressively as CheY concentration rises.

To further confirm that CheY overexpression promotes aggregation through increased c-di-GMP levels, we performed additional experiments co-overexpressing CheY and a phosphodiesterase (PDE) from E. coli to reduce intracellular c-di-GMP. These experiments showed that PDE expression mitigates cell aggregation caused by CheY overexpression (Fig. S8).

We have revised the manuscript accordingly (lines 290-294) and added these new results in Fig. S8.

Reviewer #3 (Public Review):

Summary:

The authors investigated the assembly and polar localization of the chemosensory cluster in P. aeruginosa. They discovered that a certain protein (FlhF) is required for the polar localization of the chemosensory cluster while a fully-assembled motor is necessary for the assembly of the cluster. They found that flagella and chemosensory clusters always co-localize in the cell; either at the cell pole in wild-type cells or randomly-located in the cell in FlhF mutant cells. They hypothesize that this co-localization is required to keep the level of another protein (CheY-P), which controls motor switching, at low levels as the presence of high levels of this protein (if the flagella and chemosensory clusters were not co-localized) is associated with high-levels of c-di-GMP and cell aggregations.

Strengths:

The manuscript is clearly written and straightforward. The authors applied multiple techniques to study the bacterial motility system including fluorescence light microscopy and gene editing. In general, the work enhances our understanding of the subtlety of interaction between the chemosensory cluster and the flagellar motor to regulate cell motility.

Weaknesses:

The major weakness in this paper is that the authors never discussed how the flagellar gene expression is controlled in P. aeruginosa. For example, in E. coli there is a transcriptional hierarchy for the flagellar genes (early, middle, and late genes, see Chilcott and Hughes, 2000). Similarly, Campylobacter and Helicobacter have a different regulatory cascade for their flagellar genes (See Lertsethtakarn, Ottemann, and Hendrixson, 2011). How does the expression of flagellar genes in P. aeruginosa compare to other species? How many classes are there for these genes? Is there a hierarchy in their expression and how does this affect the results of the FliF and FliG mutants? In other words, if FliF and FliG are in class I (as in E. coli) then their absence might affect the expression of other later flagellar genes in subsequent classes (i.e., chemosensory genes). Also, in both FliF and FliG mutants no assembly intermediates of the flagellar motor are present in the cell as FliG is required for the assembly of FliF (see Hiroyuki Terashima et al. 2020, Kaplan et al. 2019, Kaplan et al. 2022). It could be argued that when the motor is not assembled then this will affect the expression of the other genes (e.g., those of the chemosensory cluster) which might play a role in the decreased level of chemosensory clusters the authors find in these mutants.

We thank the reviewer for the insightful comments. P. aeruginosa possesses a four-tiered transcriptional regulatory hierarchy controlling flagellar biogenesis. Within this system, fliF and fliG belong to class II genes and are regulated by the master regulator FleQ. In contrast, chemotaxis-related genes such as cheA and cheW are regulated by intracellular free FliA, and currently, there is no evidence that FliA activity is influenced by proteins like FliG.

To verify that the expression of core chemotaxis proteins was not affected by deletion of fliG, we performed Western blot analyses to compare CheY levels in wild-type, ΔfliF, and ΔfliG strains. We observed no significant differences, indicating that the reduced presence of receptor clusters in these mutants is not due to altered expression of chemotaxis proteins.

Accordingly, we have revised the manuscript (lines 341-348) and updated Fig. 3B to reflect these findings.

Recommendations for the authors:

Reviewing Editor (Recommendations For The Authors):

The reviewers comment on several important aspects that should be addressed, namely: the lack of statistical analysis; the need for clarifications regarding assumptions made regarding receptor localization; the functional importance of receptor-motor colocalization; and the need for an elaborate discussion of flagellar gene expression. Also, two reviewers pointed out the need to prove the co-localization of CheY and CheA; This is important since CheY is dynamic, shuttling back and forth from the chemotaxis complex to the base of the flagella, whereas CheA (or cheW or, even better, the receptors) is considered less dynamic and an integral part of the chemotaxis complex.

Reviewer #1 (Recommendations For The Authors):

Minor points:

Line 43: "ubiquitous" - I would choose another word.

We changed "ubiquitous" to "widespread".

Line 49: "order" - change to organize.

We changed "order" to "organize".

Line 52: "To grow and colonize within the host, bacteria have evolved a mechanism for migrating...". Motility "towards more favorable environments" is an important survival strategy of bacteria in various ecological niches, not only within the host.

We revised it to "grow and colonize in various ecological niches".

Line 72: Define F6 in "F6 pathway-related receptors".

The proteins encoded by chemotaxis-related genes collectively constitute the F6 pathway, which we have now explained in the manuscript text.

Line 72-73: Do references 17 &18 really report colocalization of the chemotaxis receptor and flagella to the same pole? If these or other reports document such colocalization, then the sentence in the Abstract "Surprisingly, we found that both are located at the same cell pole..." is not correct.

Kazunobu et al. (ref. 18) used scanning electron microscopy to preliminarily characterize the flagellation pattern of Pseudomonas aeruginosa during cell division, showing that existing flagella are located at the old pole. Zehra et al. (ref. 17), through fluorescence microscopy, observed that CheA and CheY proteins in dividing cells are typically also present at the old pole. Based on these observations, we inferred in the Introduction that the chemotaxis complex and flagellum may localize to the same cell pole.

However, this inference is indirect and lacks direct live-cell evidence of colocalization, leaving its validity to be confirmed. This uncertainty was indeed the starting point and motivation for our study.

In our work, we simultaneously visualized flagellar filaments and core chemoreceptor proteins at the single-cell level in P. aeruginosa. We characterized the assembly and spatial coordination of the chemotaxis network and flagellar motor throughout the cell cycle, providing direct evidence of their colocalization and coordinated assembly. This represents a significant advance beyond prior indirect observations and supports the novelty of our study.

Accordingly, we have revised the relevant statements in lines 71-75 of the manuscript to better reflect the current state of the literature and emphasize the novelty of our direct observations.

Line 108: "CheY has been shown to colocalize with chemoreceptors". The authors rely here (reference 29) and in other places on findings in E. coli. However, in the Introduction, they describe the many differences between the motility systems of P. aeruginosa and E. coli, e.g., the number of chemosensory systems and their spatial distribution (E. coli is a peritrichous bacterium, as opposed to the monotrichous bacterium P. aeruginosa). There seem to be proofs for colocalization of the Che and MCP proteins in P. aeruginosa, which should be cited here.

Thank you for pointing this out. Harwood's group reported that a cheY-YFP fusion strain exhibited bright fluorescent spots at the cell pole, which disappeared upon knockout of cheA or cheW-genes encoding structural proteins of the chemotaxis complex. This strongly suggests colocalization of CheY with MCP proteins in P. aeruginosa. We have now cited this study as reference 17 in the manuscript.

Figure 1B: Please replace the order of the schematic presentations, so that the cheY-egfp fusion, which is described first in the text, is at the top.

We have modified the order of related images in Fig. 1B.

Line 127: "by introducing cysteine mutations". Replace either by "by introducing cysteines" or by "by substituting several residues with cysteines".

We changed the relevant statement to "by introducing cysteines".

Line 144-145: "Given that the physiological and physical environments of both cell poles are nearly identical.". I think that also the physical, but certainly the physiological environment of the two poles is not identical. First, one is an old pole, and the other a new pole. Second, many proteins and RNAs were detected mainly or only in one of the poles of rod-shaped Gram-negative bacteria that are regarded as symmetrically dividing. Although my intuition is that the authors are correct in assuming that "it is unlikely that the unipolar distribution of the chemoreceptor array can be attributed to passive regulatory factors", relating it to the (false) identity between the poles is incorrect.

We thank the reviewer for this important correction. We agree that the physiological environments of the two poles are not identical, given that one is the old pole and the other the new pole, and that many proteins and RNAs show polar localization in rod-shaped Gram-negative bacteria. Accordingly, we have revised the original text (lines 150-152) to read:

“Despite potential differences in the physical and especially physiological environments at the two cell poles, it is unlikely that the unipolar distribution of the chemotaxis complex can be attributed to passive regulatory factors.”

Lines 151-154: "Considering the consistent colocalization pattern between chemosensory arrays and flagellar motors in P. aeruginosa". Does the word consistent relate to different reports on such colocalization or to the results in Figure 1D? In case it is the latter, then what is the word consistent based on? All together only 7 cells are presented in the 5 micrographs that compose Figure 1D (back to statistics...).

We thank the reviewer for raising this point. To clarify, the word "consistent" refers to the observation of colocalization shown in Figure 1D & Figure S3. As noted in the revised figure legend for Figure 1D, a total of 145 cells with labeled flagella were analyzed, all exhibiting consistent colocalization between flagella and chemosensory arrays. Additionally, we have included a new image showing a large field of co-localization in the wild-type strain as Figure S3 to better illustrate this consistency.

Figure 2A: Omit "Subcellular localization of" from the beginning of the caption.

We removed the relevant expression from the caption.

Reviewer #2 (Recommendations For The Authors):

I strongly recommend checking that CheY localizes to the receptor cluster in PA. This could be done by tagging cheA with a different fluorophore and demonstrating their colocalization. It would also be helpful to check that they are colocalized in the delta flhF mutant.

We thank the reviewer for this valuable suggestion. We constructed a plasmid expressing CheA-CFP and introduced it into the CheY-EYFP strain by electroporation. Fluorescence imaging revealed a high degree of spatial overlap between CheA-CFP and CheY-EYFP signals, indicating that CheY-EYFP indeed marks the location of the chemoreceptor complex.

We have revised the manuscript accordingly (lines 118-123) and included these results in the new Fig. S2.

The experiments under- and over-expressing CheY part seemed too unrelated to receptor-motor colocalization. I think the authors should think about a more direct way of testing whether colocalization of the motor and receptors is important for preventing signaling crosstalk. One way would be to measure cdG levels in WT and in delta flhF mutants and see if there is a significant difference.

We thank the reviewer for raising the important point regarding potential cellular stress caused by elevated CheY concentrations, as well as for the suggestion to test the hypothesis using flhF mutants.

First, as noted in the response to your 2nd comment in Public Review, CheY-P concentration rapidly decreases away from the receptor complex. While deletion of flhF alters the position of the receptor complex, thereby shifting the region of high CheY-P concentration, it does not increase CheY-P levels elsewhere in the cell. Importantly, in the ΔflhF strain, the receptor complex and the motor still colocalize, so this mutant may not effectively test the role of receptor-motor colocalization in preventing crosstalk as suggested.

Regarding the possibility that elevated CheY levels stress the cells independently of CheY-P signaling, prior work in E. coli by Cluzel et al. (ref. 11) showed that overexpressing CheY several-fold did not cause phenotypic changes, indicating that simple CheY overexpression alone may not be generally stressful. Furthermore, our data indicate that the increase in c-di-GMP levels and subsequent cell aggregation upon CheY overexpression is not an all-or-none switch but occurs progressively as CheY concentration rises.

To further confirm that CheY overexpression promotes aggregation through increased c-di-GMP levels, we performed additional experiments co-overexpressing CheY and a phosphodiesterase (PDE) from E. coli to reduce intracellular c-di-GMP. These experiments showed that PDE expression mitigates cell aggregation caused by CheY overexpression (Fig. S8).

We have revised the manuscript accordingly (lines 290-294) and added these new results in Fig. S8.

Reviewer #3 (Recommendations For The Authors):

(1) Can the authors elaborate more on the hierarchy of flagellar gene expression in P. aeruginosa and how this relates to their work?

We thank the reviewer for the suggestion. We have now described the hierarchy of flagellar gene expression in P. aeruginosa in lines 341-348.

(2) I would suggest that the authors check other flagellar mutants (than FliF and FliG) where the motor is partially assembled (e.g., any of the rod proteins or the P-ring protein), together with FlhF mutant, to see how a partially assembled motor affects the assembly of the chemosensory cluster.

We thank the reviewer for this valuable suggestion. The P ring, primarily composed of FlgI, acts as a bushing for the peptidoglycan layer, and its absence leads to partial motor assembly. We constructed a ΔflgI mutant and observed that the proportion of cells exhibiting distinct chemotactic complexes was similar to that of the wild-type strain, suggesting that the assembly of the receptor complex is likely influenced mainly by the C-ring and MS-ring structures rather than by the P ring. We have revised the original text accordingly (lines 217-220) and added the corresponding data as Figure S6.

(3) I would suggest that the authors check the levels of CheY in cells induced with different concentrations of arabinose (i.e., using western blotting just like they did in Figure 3B).

We have assessed the levels of CheY in cells induced with different concentrations of arabinose using western blotting, as suggested. The results have been incorporated into the manuscript (lines 274-275) and are presented in Figure S7.

(4) To my eyes, most of the foci in FliF-FlhF mutant in Figure 3A are located at the pole (which is unlike the FlhF mutant in Figure 2). Is this correct? I would suggest that the authors also investigate this to see where the chemosensory cluster is located.

We thank the reviewer for pointing this out. The distribution of the chemotaxis complex in the ΔflhFΔfliF strain was investigated and showed in Fig. S4. Indeed, most of the chemoreceptor foci in this mutant are located at the pole. This probably suggests that, in the absence of both FlhF and an assembled motor, the position of the receptor complex may be largely influenced by passive factors such as membrane curvature. This interesting possibility warrants further investigation in future studies.

Version 2

Abstract

Abstract

The opportunistic pathogen Pseudomonas aeruginosa serves as a model organism for studying multiple signal transduction pathways. The chemoreceptor cluster, a core component of the chemotaxis pathway, is assembled from hundreds of proteins. The unipolar distribution of receptor clusters has long been recognized, yet the precise mechanism governing their assembly remains elusive. Here, we directly observed the relative positions of the flagellar motor and chemoreceptor cluster using flagellar filament labeling and gene editing techniques. Surprisingly, we found that both are located at the same cell pole, with the distribution pattern controlled by the polar anchor protein FlhF. Additionally, the efficient assembly of the chemoreceptor cluster is partially dependent on the integrity of the motor structure. Furthermore, we discovered that overexpression of the chemotaxis regulatory protein CheY leads to high intracellular levels of the second messenger c-di-GMP, triggering cell aggregation. Therefore, the colocalization of the chemoreceptor cluster and flagellum in P. aeruginosa serves to avoid cross-pathway signaling interference, enabling cells to conduct various physiological activities in an orderly manner.

Introduction

In biological systems, the process by which proteins self-assemble into organized complex structures is widespread1,2. Pole-to-pole protein oscillations in the Min system ensure that the FtsZ ring, a crucial component of cell division, is placed precisely in the middle of the cell body3,4. In response to antibiotic treatment and heat stress, some bacteria generate multiple protein foci throughout the cytoplasm5,6. The polar localization of the secretion system, such as the type VI secretion system, is mediated by targeting proteins and potentially facilitates host-pathogen interactions7,8. Thus, spontaneous assembly is accompanied by the formation of spatial patterns that organize the intracellular environment.

To grow and colonize in various ecological niches, bacteria have evolved a mechanism for migrating towards more favorable environments. The locomotion organ (the flagellar motor) and the chemotaxis pathway play crucial roles in achieving this goal. The former provides physical drive, while the latter offers directional guidance. The initiation of cell reorientation is controlled by motor switching, which is governed by the chemotaxis two-component signaling pathway. This pathway comprises transmembrane chemoreceptors, a histidine kinase CheA, an adaptor protein CheW, a response regulator CheY, and two adaptation proteins, CheB and CheR. Receptors collaborate with CheA and CheW to form chemotaxis complex in a hexagonal pattern9,10. Upon phosphorylation of CheY by the kinase at the complex, it freely diffuses to the motor and binds to FliM, a component of the motor switch complex, to affect motor switching11–13. It is worth noting that there exists diversity among different bacterial species in the placement of flagellar motors14, whereas chemoreceptor clusters are typically found at one or both ends of the cell15. These two structural units are functionally connected. However, it remains unclear whether there is an interaction in their spatial distribution and what the specific regulatory mechanism might be.

Pseudomonas aeruginosa, a common human opportunistic pathogen, possesses four chemosensory pathways that perform distinct functions and are stimulated by signal binding to 26 chemoreceptors16. Among them, the proteins encoded by chemotaxis-related genes collectively constitute the F6 pathway. Previous studies have suggested that the receptors involved in this pathway localize to the old cell pole17, similar to the flagellum18, implying colocalization of the receptors and the flagellum at the same pole; however, direct evidence for this colocalization is still lacking. Furthermore, the polar anchor protein FlhF has been reported to guide the flagellum to grow at the cell pole19,20. However, the mechanism by which chemotaxis complex-related proteins aggregate and self-assemble at the cell pole is controversial. The distribution of the chemotaxis complex in the peritrichous bacterium Escherichia coli has been extensively studied and attributed to several mechanisms, including stochastic self-assembly21,22, membrane curvature sorting23,24, and inefficient clustering in the lateral region25. However, these mechanisms cannot explain the unipolar distribution pattern observed in P. aeruginosa. Unlike E. coli, specific genes responsible for the placement of the chemotaxis complex have been identified in several bacterial species. In Caulobacter crescentus, chemotaxis complex assemble at the new cell pole with the help of TipN and TipF proteins26. A tripartite ParC-ParP-CheA interaction network was reported to promote polar localization of chemotaxis complex in Vibrio parahaemolyticus27,28. However, as there are no related genes in P. aeruginosa, the mechanism of its chemotaxis complex distribution remains a mystery.

Here, we combined gene editing and in vivo fluorescence imaging of flagellar filaments to directly observe the distribution of chemotaxis complex and flagellar motors, proposing a cooperative construction model of chemotaxis network and flagella during the entire division cycle of P. aeruginosa. The core focus of this study is to clarify the regulatory mechanism of its chemotaxis complex distribution. We found a substantial association between the assembly of the flagellar motor and the chemotaxis complex. The complete assembly of the motor serves as a partial prerequisite for the assembly of the chemotaxis complex, and its assembly site is also regulated by the polar anchor protein FlhF. Furthermore, by introducing exogenously expressed CheY protein, we found that this triggers the expression of c-di-GMP at a high level. From this, we infer that the colocalization of the chemotaxis complex and the flagellum in P. aeruginosa avoids the cross-pathway interference of signaling molecules, thus providing a guarantee for the coexistence of multiple chemosensory pathways.

Results

Robust generation of daughter cell with both chemotaxis network and flagellar motor

CheY has been shown to colocalize with chemoreceptors17,29. To visualize the distribution of chemotaxis complex in cells, we fused the gene encoding the enhanced yellow fluorescent protein (eYFP) to the cheY gene in the P. aeruginosa chromosome (Fig. 1A). All genetic modifications were carried out at the native position on the chromosome, allowing the expression of chemotaxis proteins in precise stoichiometry from their natural promoters (Fig. 1B). The mutant strain with cheY-eyfp was able to moderately expand on soft agar plates17, proving that its chemotaxis behavior was not notably affected, so it was selected for subsequent experiments. As shown in Fig. 1C, CheY-eYFP was mainly located at the cell pole, suggesting that CheA, CheW and CheY in P. aeruginosa form a signal transduction complex that was primarily distributed at the cell pole, as described previously17. Representative large-field images containing many cells are shown in Fig. S1. Additionally, we constructed a plasmid expressing CheA-CFP and electroporated it into the cheY-eyfp strain. Fluorescence imaging revealed a high degree of spatial overlap between CheA-CFP and CheY-EYFP (Fig. S2), confirming that CheY-EYFP accurately marks the location of the chemoreceptor complex. From our measurements, nearly 90% (332/372) of cells contain obvious receptor clusters, and factors such as fluorescence bleaching may have caused the loss of fluorescent spots in some cells. In addition, we observed the presence of fluorescent spots at both ends of some cells with a large aspect ratio (probably approaching cell division), indicating that the newly generated progeny cells will have a complete chemotaxis network. Next, we sought to observe the distribution of the flagella in cells simultaneously to understand the motility of future progeny.

A. Schematic diagram of the chemotaxis signal transduction network and flagellar motor of P. aeruginosa. Fusion protein CheY-EYFP and fluorescently labeled flagellar filaments were used as markers to indicate the position of the chemotaxis complex and motor, respectively. B. The labeling mechanism of flagellar filaments and chemotaxis regulatory protein CheY. Filaments (with cysteine point mutation FliCT394C) were labeled through sulfhydryl-maleimide conjugation, and cheY-eyfp fusion with a 3× glycine linker was used to visualize chemotaxis complex positions. C. Localization of CheY-EYFP in the wild-type strain of P. aeruginosa. CheY-EYFP is mainly located at the single cell pole, and the white arrow points to individuals with obvious chemotaxis complex at both cell poles, which generally have a large aspect ratio of the cell body. The yellow dashed box marks the cell outline. D. The merged imaging of flagellar filaments and CheY-EYFP in the wild-type strain of P. aeruginosa, where flagellar motor and chemotaxis complex colocalize in cells. 145 cells with labeled flagella were observed, all of which exhibited consistent colocalization. White arrows point to individuals about to be divided, and the yellow dashed box marks the cell outline. The scale bar is 1 μm.

Our recent study demonstrated the successful labeling of the flagellar filament with thiol-reactive fluorescent dye by introducing cysteines into the flagellin FliC30,31 (Fig. 1B). To avoid fluorescence interference, we utilized a dye with an excitation peak near 568 nm, considering that the excitation peak of eYFP is 513 nm, and the emission peak is 527 nm. We employed a xenon lamp as the excitation light source and switched filters to enable simultaneous fluorescence imaging of flagellar filaments and chemotaxis complex within the same cell.

Remarkably, our findings revealed a surprising colocalization of chemotaxis complex and flagellar filaments at the same end of the cell body, and this colocalization was consistent, as shown in Fig. 1D. A typical large field was shown in Fig. S3. Similar to the chemotaxis complex, we observed the phenomenon of double flagella symbiosis in cells with a large aspect ratio, suggesting that P. aeruginosa has evolved a cell division mechanism with precise timing regulation. Before cell division, both poles of the mother cell assembled chemotaxis complex and flagellar motors. This ensures that daughter cells possess complete motility and chemotaxis, thereby greatly enhancing the environmental adaptability of the population.

FlhF mediates proper unipolar localization of the chemotaxis complex

Despite potential differences in the physical and especially physiological environments at the two cell poles, it is unlikely that the unipolar distribution of the chemotaxis complex can be attributed to passive regulatory factors. The number and distribution of flagellar motors and chemotaxis complex vary among different bacterial species14,15, and the localization system for flagella has been well studied. Specifically, the FlhF-FlhG system, discovered in several monotrichous bacterial species, has been shown to control the location and number of flagella32,33. In P. aeruginosa, a knockout of flhF leads to mis-localized flagellar assembly19,20. Considering the consistent colocalization pattern between chemotaxis complex and flagellar motors in P. aeruginosa, we speculate that the distribution of chemotaxis complex is also regulated by similar molecular mechanisms.

To investigate the role of FlhF in the localization of chemotaxis complex, we constructed a ΔflhF strain for fluorescence observation. The results revealed that the chemotaxis complex no longer grow robustly at the cell pole (Fig. 2A), and the assembly positions of the flagellar motor change accordingly, colocalizing with the chemotaxis complex (Fig. 2B). Representative large fields containing many cells are shown in Fig. S1. Additionally, we categorized receptor cluster distribution patterns into three types: precise polar, near polar, and mid-cell. The relevant statistics for the wild-type strain and the ΔflhF mutant are presented in Fig. S4, which show a significant increase in the proportion of non-precise polar chemoreceptors following flhF knockout. We also quantified the proportion of individuals with obvious receptor clusters, which was more than 80% (182/221), similar to the wild-type strain. These experimental results suggest that the polar anchoring protein FlhF has a minimal impact on the assembly efficiency of P. aeruginosa’s chemotaxis complex, but it strictly controls its unipolar distribution. Furthermore, a consistent colocalization between the flagellar motor and chemotaxis complex was observed, independent of FlhF. Thus, it is crucial to determine whether there is a causal relationship in the assembly order of these two structural units.

A. Localization of CheY-EYFP in the ΔflhF strain of P. aeruginosa. CheY-EYFP is no longer robustly distributed at the single cell pole. The yellow dashed box marks the cell outline. B. The merged imaging of flagellar filaments and CheY-EYFP in the ΔflhF strain of P. aeruginosa, flagellar motor and chemotaxis complex still colocalize in cells. 101 cells with labeled flagella were observed, all of which exhibited consistent colocalization. The yellow dashed box marks the cell outline. The scale bar is 1 μm.

Motor structural integrity is a prerequisite for chemoreceptor self-assembly

Cryo-electron microscopy has successfully revealed the complete flagella structure in various bacterial species. This structure includes multiple components such as the inner-membrane MS ring, the cytoplasmic C ring, and the internal protein secretion system, all of which exhibit strong similarities34. This suggests that the core structure of the flagellar motor is highly conserved. The MS ring, composed of 26 FliF proteins, forms the foundation of the flagellar structure35. In addition to serving as a mounting platform for the C ring, the MS ring also acts as a protective shell for the internal protein secretion machinery36. FliG, the main component protein of the C ring, binds directly to the cytoplasmic surface of FliF in a 1:1 ratio37. Thus, the efficient assembly of the C ring requires preassembly of the MS ring. To observe the distribution of the chemotaxis complex following the disruption of the C ring and MS ring, we constructed ΔfliG and ΔfliF mutants.

We performed fluorescence observation on the ΔfliG mutant, which has a disrupted flagellar motor C ring. The results showed that the proportion of ΔfliG cells with obvious chemoreceptor clusters decreased significantly compared to the wild-type cells (59.8%, 140/234), although the chemotaxis complex remained at a single cell pole. We also examined the ΔfliF mutant, which has a disrupted flagellar motor MS ring. Similar to the ΔfliG mutant, the proportion of individuals with obvious chemotaxis complex decreased significantly (62.5%, 202/323). Subsequently, we constructed a ΔflhFΔfliF mutant and observed that the proportion of individuals with obvious chemoreceptor clusters further decreased (50.7%, 341/672). To ascertain whether it is motor integrity rather than functionality that influences the efficiency of chemotaxis complex assembly, we constructed a stator mutant (ΔmotAΔmotCD). In this mutant, the motor is completely stalled while the structure remains intact. We found that the mutant performed similarly to the wild-type strain in terms of chemotaxis complex assembly (84.3% of individuals with obvious clusters, 204/242). The fluorescence imaging of receptor clusters for multiple mutants in this section is shown in Figure. 3A, and corresponding large-field images are provided in Fig. S1. We utilized Western blotting to measure the expression levels of CheY, which were found to be similar across the different strains (Fig. 3B). These further substantiated that the observed phenomenon is based on the structural integrity of the motor rather than the protein expression level. In contrast, the cheA knockout strain was constructed to disrupt the assembly of the chemoreceptor complex. We fluorescently labeled its flagellar filaments and found that its phenotype was no different from that of the wild-type strain (Fig. S5), indicating that the assembly of the chemotactic receptor complex does not affect the flagellar assembly efficiency. The P ring (mainly composed of FlgI) is thought to act as a bushing for the peptidoglycan layer, and its absence results in partial assembly of the motor structure38. We constructed ΔflgI mutant and found that the proportion of cells with distinct chemotactic complexes was similar to that of the wild-type (Fig. S6). Additionally, we quantified the proportion of cells with receptor clusters (Fig. 3C). Overall, our findings suggest that the polar anchor protein FlhF and the structural integrity of the motor are crucial for the formation of chemotaxis complex. The former guides them to the appropriate site, while the latter influences the assembly efficiency of the chemoreceptor complex.

A. Localization of CheY-EYFP in various P. aeruginosa strains. The scale bar is 10 μm. B. Western blot analysis was performed to detect CheY expression in various P. aeruginosa strains. β-actin was used as the housekeeping protein. C. Occurrence probability of obvious chemotaxis complex in P. aeruginosa wild-type and several mutant strains. The proportion of individuals with obvious chemotaxis complex decreased significantly in the motor-incomplete strains (ΔfliF and ΔfliG), and this value was further reduced after flhF knockout (ΔfliFΔflhF). The ΔfliF and ΔmotAΔmotCD strains have a similar chemotaxis complex occurrence probability as the wild-type strain. The number of cells analyzed for each strain (from left to right) were: 372, 221, 234, 323, 672, and 242. ‘***’: significant difference (P-value<0.0001), ‘ns.’: no significant difference (P-value>0.05).

Colocalization of chemotaxis complex and flagellar motor avoids cross-pathway interference in P. aeruginosa signal transduction

The distribution patterns of receptors associated with various signal transduction pathways in P. aeruginosa vary considerably. For instance, the biofilm formation-related receptor WspA, unlike the chemotaxis receptors discussed here, is distributed throughout the cell. This distribution is thought to enhance its sensitivity to mechanical perturbations of the cell membrane39. Previous studies speculated that the highly consistent position of chemotaxis complex and flagellar motors enhances bacterial chemotaxis performance28. However, the diffusion time of the phosphorylated chemotaxis regulatory protein CheY-P across the longest distance in a bacterial cell body (along the cell’s long axis) is approximately 100 ms40,41, whereas the timescale for the chemotaxis temporal comparison is on the order of seconds42. Additionally, a study by Fukuoka and colleagues reported that intracellular chemotaxis signal transduction requires approximately 240 ms beyond CheY or CheY-P diffusion time41. Moreover, the CCW/CW interval of the P. aeruginosa flagellar motor under normal conditions is 1-2 s, as determined by bead assays30 or tethered cell assays43. Taken together, these indicate that for P. aeruginosa, which moves via a run-reverse mode, the potential 100 ms reduction in response time due to co-localization of the chemotaxis complex and motor has a limited effect on overall chemotaxis timing. Consequently, the physiological significance of the colocalization of chemoreceptors and flagellar motors remains unclear.

E. coli mediates chemotaxis through a single pathway involving five types of chemoreceptors44. As a core regulatory protein for chemotaxis, CheY participates in multiple processes such as adaptation and chemotaxis response, with CheY-P binds to the motor to regulate switching11. In contrast, P. aeruginosa has four distinct chemosensory pathways that perform different functions, involving different CheY homologs, and are stimulated by signals binding to 26 types of chemoreceptors. Consequently, P. aeruginosa possesses more complex chemosensory pathways16. We thus hypothesized that the co-localization of chemoreceptors and flagellar motors in P. aeruginosa ensures locally distributed CheY-P molecules, thereby eliminating the need for a high level of intracellular CheY-P and avoiding potential side effects on other signaling pathways.

To test the potential effect of an increased intracellular CheY-P level, we constructed the CheY expression plasmid cheY-pJN105, transformed it into the wild-type P. aeruginosa strain, and induced it with varying concentrations of arabinose. We observed that the higher the inducer concentration, the more pronounced the cell aggregation became in the field of view (Fig. 4A). The transition from planktonic individuals to aggregated communities is similar to biofilm formation, which is known to be accompanied by a significant increase in intracellular c-di-GMP levels.

A. The evolution of cell aggregation as the intracellular CheY concentration increases by induction with higher concentrations of arabinose. The scale bar is 10 μm. B. Quantitative characterization of intracellular c-di-GMP levels at different CheY concentrations. From top to bottom, they correspond to the ΔcheY strain (N=198), wildtype strain (N=249) and CheY overexpression strain (N=228), respectively.

To investigate whether a higher level of intracellular CheY-P increased the intracellular c-di-GMP level, we sought to detect the c-di-GMP level. We introduced the plasmid pCdrA-gfp into P. aeruginosa, using it as a c-di-GMP biosensor with the fluorescence intensity proportional to the c-di-GMP level45. To ensure the coexistence of the plasmids, the intracellular CheY concentration was controlled by the plasmid cheY-pME6032. The intracellular CheY levels induced by varying arabinose concentrations were measured, showing clear differences across the concentration gradient (Fig. S7). We found that the average single-cell fluorescence intensity of the CheY overexpression strain was 71.8% higher than that of the wild-type strain, while the c-di-GMP level of the cheY deletion strain was 58.6% lower than that of the wild-type (Fig. 4B).

The colocalization of both the signaling source (the kinase) and sink (the phosphatase) at the chemoreceptor complex at the cell pole results in a rapid decay of CheY-P concentration within approximately 0.2 µm from the cell pole, leading to a nearly uniform distribution elsewhere in the cell, as demonstrated by Vaknin and Berg46. This spatial arrangement effectively confines high CheY-P levels to the pole region. When the motor is also localized at the cell pole, the need for elevated CheY-P concentrations throughout the cytoplasm is reduced. Therefore, the colocalization of chemotaxis complex and flagellar motor facilitates the precise regulation of cell swimming direction within the low intracellular CheY-P concentration threshold, using locally distributed CheY-P molecules. This helps to avoid the occurrence of the aforementioned cross-pathway interference. To further confirm that CheY overexpression promotes aggregation through increased c-di-GMP levels, we performed additional experiments co-overexpressing CheY and a phosphodiesterase (PDE) from E. coli to reduce intracellular c-di-GMP. These experiments showed that PDE expression mitigates cell aggregation caused by CheY overexpression in P. aeruginosa (Fig. S8).

Summary and discussion

P. aeruginosa harbors multiple signal transduction systems that regulate flagella-mediated swimming motility (Che pathway), pili-mediated interface twitching motility (Pil pathway), and biofilm formation (Wsp pathway)16. These systems allow it to thrive in various ecological niches within complex external environments. Here, employing a combination of chromosomal fluorescent protein fusion and flagellar labeling techniques in living cells, we directly observed that the chemotaxis complex of P. aeruginosa Che pathway and the flagellar motor share a high degree of consistency in their assembly sites. We found that the chemotaxis complex, comprising Che proteins, was consistently located at the cell pole where the flagellum was positioned. Based on these observations, we deduced the construction mode of the chemotactic network and flagellar motor during the entire cell growth cycle (Fig. 5A). As the cell body matures and elongates, a new flagellum grows at the opposing cell pole, accompanied by the assembly of a fresh chemotaxis complex. The cell then undergoes division from the middle, generating two daughter cells, each equipped with fully functional motility and chemotaxis capabilities. This ensures robust inheritance of these chemotaxis and motility-related macromolecular machines.

A. The construction mode of the chemotaxis network and flagellar motor throughout the complete cell growth cycle. B. The proximal growth mode of the P. aeruginosa flagellar motor and receptor clusters will effectively regulate the spatial range of CheY action, thus avoiding unintentional cross-pathway regulation.

Based on our understanding of the regulation mechanism of flagellar position in P. aeruginosa, we constructed a ΔflhF mutant and found that the position distribution of its chemotaxis complex is actively regulated by FlhF-mediated molecular mechanisms, rather than being passively restricted by membrane curvature factors as seen in E. coli. A similar active regulation phenomenon was previously reported in Vibrio cholerae, with the corresponding molecular regulation mechanism detailed extensively28. The flagella and chemotaxis complex of newborn V. cholerae cells are located at the old cell pole, with the polar anchor protein HubP distributed at both cell poles. As the cell grows, HubP recruits the ParA homologue ParC. Subsequently, ParC recruits ParP to assemble a new chemotaxis complex at the new cell pole. Upon cell division, HubP relocates to the middle of the cell, ensuring its robust presences at both poles of the daughter cells. Here, HubP recruits ParA homologues including FlhG, which influence flagellar motor assembly. However, previous experiments confirmed that the cell cycle-dependent polar localization of ParC and chemotactic proteins is independent of the flagellar position regulatory protein FlhF. This suggests that the localization of flagellar motors and chemotaxis complex in V. cholerae is controlled separately through different pathways. In contrast, flagellar motors were no longer robustly distributed at the cell pole after flhF was knocked out in P. aeruginosa, whereas the assembly site of chemotaxis complex and flagellar motor remained remarkably consistent. This suggests that FlhF regulates the distribution patterns of both and implies an interaction between them.

We further constructed ΔfliG and ΔfliF mutants, which disrupted the assembly of the flagellar motor. Surprisingly, we observed that the assembly of chemotaxis complex was impaired under these conditions, with a significant decrease in the proportion of individuals exhibiting obvious Che protein fluorescent bright spots. In contrast, the chemotaxis complex assembly in the stator mutant (ΔmotAΔmotCD) was similar to that of the wild-type strain. It should be noted that P. aeruginosa has a four-tiered transcriptional regulatory circuit controlling flagellar biogenesis, in which FliG and FliF belong to class II and are regulated by the master transcriptional regulator FleQ47.

Chemotaxis complex-related genes (cheA, cheW, etc.) are regulated by intracellular free FliA proteins, and there is no evidence that FliA is regulated by proteins such as FliG. To further confirm that the expression of core chemotaxis proteins was not affected by the knockout of fliG, we performed Western blotting to measure CheY expression levels in wild-type, ΔfliF and ΔfliG strains, and found no significant differences (Fig. 3B). These results suggested that the integrity rather than functionality of the flagellar motor in P. aeruginosa is critical for the assembly of its chemotaxis complex, and its assembly efficiency is greatly reduced in the absence of flagellar motors. To eliminate the potential confounding effects of FlhF and flagellar motors, we constructed a ΔflhFΔfliF mutant strain. Under this condition, the proportion of individuals with obvious Che protein fluorescent bright spots dropped to ∼50%, indicating that passive regulation based on membrane curvature may still exist in P. aeruginosa, albeit to a lesser degree.

Is the remarkable consistency in the spatial distribution of these two independent structural units an unintentional outcome or a hidden mystery? Given that Kulasekara et al. previously proposed that CheA and its phosphorylation state can cause c-di-GMP heterogeneity among individual P. aeruginosa cells through a specific PDE48,49, we sought to verify whether the phenotypic differences induced by CheY here are closely related to changes in the overall population level of c-di-GMP. We introduced a c-di-GMP monitoring system at the single-cell level and discovered that overexpression CheY substantially increases the intracellular c-di-GMP concentration. Previous studies generally believed that CheY-P, as a chemotactic regulatory protein, only influences flagellar motor switching. Here, we found that CheY-P, as a signaling molecule, can accomplish threshold-limited cross-pathway regulation. At low CheY-P concentrations, it is conservatively involved in motor switching regulation, as confirmed by previous studies. At high CheY-P concentrations, cell motility is suppressed, and c-di-GMP is triggered to express, though its specific mechanism needs further exploration. Unlike CheA, a structural protein of the chemotaxis complex, CheY is a protein that can shuttle in the cytoplasm and can also regulate the intracellular c-di-GMP level. This finding expands our understanding of the connection between the internal chemotaxis network and the c-di-GMP pathway. The intracellular environment is highly crowded50, and the proximal growth pattern shown in Fig. 5B has multiple physiological significances. On one hand, chemotaxis-related protein can precisely regulate cell motility with minimal synthesis costs. On the other hand, the chemotactic network and the flagellar motor are spatially integrated within the cell, ensuring their robust existence within the cell body and preventing mutual interference with collateral pathways.

Many bacteria possess multiple signal transduction pathways. The orderly operation of each pathway within the micrometer-scale space presents a fascinating scientific problem. Here, we utilized the well-known chemotaxis network as a starting point to shed light on this problem, providing inspiration for future in-depth understanding of related issues.

Materials and methods

Strains and cell culture

The strains and plasmids used in this study are listed in Table 1. The Escherichia coli TOP10 strain was used for standard genetic manipulations. A single-colony isolate was grown in 3 ml of LB broth (1% Bacto tryptone, 0.5% yeast extract and 1% NaCl) overnight to saturation on a rotary shaker (250 rpm) at 37°C. An aliquot was diluted 1:100 into 10 ml of LB broth and grew to exponential phase. Appropriate antibiotics were added if necessary to prevent plasmid loss: for E. coli, 15 μg/ml Gentamicin and 25 μg/ml Tetracycline; for P. aeruginosa, 30 μg/ml Gentamicin and 50 μg/ml Tetracycline. To induce protein expression, Isopropyl β-D-1-thiogalactopyranoside (IPTG, 1.0 mM) was added to strains with pME6032-derivative vectors, and arabinose (0.005%, 0.01%, 0.05%, 0.1%) was added to strains with pJN105-derivative vectors. 2 ml of cells were harvested by centrifugation at 2000×g for 2 min, washed twice in an equal volume of motility buffer (MB) [50 mM potassium phosphate, 15 μM EDTA, 0.15 M NaCl, 5 mM Mg2+ and 10 mM lactic acid (pH 7.0)]51, and resuspended in 5 ml MB for subsequent fluorescence observation.

Construction of in-frame deletion mutants

We used polymerase chain reaction (PCR) to generate ∼1000 bp DNA fragments with upstream (Up) and downstream (Dn) sequences flanking the target gene (including flhF, fliF, motA and fliG) to be deleted. The Up and Dn DNA sequences were fused with the linearized pex18gm vector using Gibson assembly52. The resulting vectors were transformed into P. aeruginosa by electroporation, and the desired knockout mutants were obtained by double selections on gentamicin plates (LB plates with 30 µg/ml gentamicin) and sucrose plates (NaCl-free LB plates with 15% sucrose) at 37°C53.

Generation of chromosomal fusions of yfp to P. aeruginosa gene

To generate eyfp fusion at the N-terminus of cheY, we first used PCR to generate upstream and downstream DNA sequences flanking the cheY gene. A subsequent PCR reaction aimed to yield the cheY-eyfp fusion with a 3× glycine linker (GGCGGAGGA), with pVS88 serving as the source of the enhanced yellow fluorescent (eyfp). The three DNA sequences, along with the linearized vector, were fused to cheY-eyfp-pex18gm using Gibson assembly. Finally, gene replacement was used to transfer the fusion constructs into the P. aeruginosa chromosome by homologous recombination.

Flagellar staining and fluorescence imaging

Flagellar filaments were labeled by following the protocol described previously54. Cells (1 ml of exponential-phase culture) were harvested by centrifugation at 2000×g for 10 min and washed twice in 1 ml MB. The final pellet was adjusted to a volume of ∼100 µl that concentrated the bacterial 10-fold. Alexa Fluor 568 maleimide (Invitrogen-Molecular Probes) was added to a final concentration of 20 μg/ml, and labeling was allowed to proceed for 30 min at room temperature with gyration at 80 rpm. Unused dye was then removed by washing the cells with MB three times, and the final pellet was resuspended in MB.

For fluorescence imaging, 50 μl of cells were added to the chamber (constructed with two double-sticky tapes, a glass slide, and a poly-L-Lysine coated coverslip), incubated for 3 min, and rinsed with 100 μl MB. The boundary of the chamber was then sealed with Apiezon vacuum grease. The chamber was placed on a Nikon Ti-E inverted fluorescence microscope with a 100× oil-immersion objective and a sCMOS camera (Primer95B, Photometrics). The flagellar filament and chemotaxis complex of cells were observed separately with corresponding filter-set, using a 200-ms exposure time. To quantify the fluorescence intensity of the receptor cluster, we took the intensity maximum point as the center and extract a 3×3 pixel matrix around it. The mean value of the elements in this matrix represents the final fluorescence intensity.

Western Blotting

The total protein of P. aeruginosa strain cells in each treatment group was extracted and its concentration was measured with Bicinchoninic acid (BCA Protein Quantification Kit, Yeasen Biotechnology Co., Ltd) when the cells grew to exponential phase with same reduced turbidity (OD600 between 0.9 and 1.0). The exact amount of protein was subjected to SDS-PAGE electrophoresis, then transferred to a nitrocellulose membrane, which was blocked with 50 mg/mL skim milk in TBST buffer (20 mM Tris, 150 mM NaCl [pH7.4], 0.1%Tween 20) for 1h at room temperature. Since the N-terminal sequence of EYFP is almost identical to that of GFP, a rabbit anti-GFP antibody (1:2000, Yeasen Biotechnology Co., Ltd) and a Mouse anti-β-actin antibody (1:2000, Invitrogen) were added and incubated at 4°C overnight. The nitrocellulose membrane was gently washed with TBST for 20 min, three times. Subsequently, HRP labeled goat anti-rabbit IgG (1:4000, Abcam) and rabbit anti-mouse IgG (1:4000, Abcam) were added and incubated at room temperature for 1 h. The nitrocellulose membrane was again gently washed with TBST for 20 min, three times. Then, the membrane was developed using the ECL chemiluminescence detection system (Beyotime P0018FS) and visualized with the ChemiDoc XRS+ system (Bio-Rad). Three independent experiments were conducted for reproducibility.

Monitoring c-di-GMP signal at the single-cell level

For measurements of c-di-GMP signaling, the validated reporter plasmid pCdrA-gfp, which produces green fluorescent protein (GFP) in response to an increase in c-di-GMP, was introduced into the experimental system. A 1000× diluted culture was inoculated into the apparatus described in the previous section and allowed to adhere to the surface for ∼10 min, with the slight difference that the coverslips were not coated with poly-L-Lysine to avoid interference from background fluorescence. Bacteria were imaged using a Nikon Ti-E inverted microscope, which was equipped with an EMCCD camera (DU897, iXon3, Andor Technology). We used a 100× oil-immersion objective. Bacteria were illuminated with a 488-nm laser (Sapphire 488−200 mW, Coherent), using a standard GFP filter sets in the fluorescence imaging system with an exposure time of 500 ms. To ensure the consistency of culture and observation conditions, the data of wild-type and cheY-pME6032 carried strains were collected at the same time. More than 200 independent individuals of each strain were used for data analysis.

Single-cell c-di-GMP concentration was characterized by calculating fluorescence intensity. The average fluorescence intensity of a single cell was calculated by dividing the total fluorescence intensity of a single cell by the cell volume. Bacterial were simplified to a hollow cylindrical trunk with hollow hemispherical caps at both ends. The cell volume can be calculated by taking the length of the short axis of the cell body as the diameter of the sphere and the length of the long axis as the sum of the height of the cylinder and the diameter of the sphere. The mean fluorescence intensity of the cell was obtained by subtracting the average background fluorescence intensity from the total fluorescence intensity of the cell, and then dividing by the cell volume.

eLife Assessment

This valuable study by Wu et al presents data on bacterial cell organization, demonstrating that the two structures that account for bacterial motility - the chemotaxis complex and the flagella - colocalize to the same pole in Pseudomonas aeruginosa cells. The work provides convincing results for the regulation underlying this spatial organization and its functioning.

Reviewer #1 (Public review):

Summary:

The study by Wu et al presents interesting data on bacterial cell organization, a field that is progressing now, mainly due to the advances in microscopy. Based mainly on fluorescence microscopy images, the authors aim to demonstrate that the two structures that account for bacterial motility, the chemotaxis complex and the flagella, colocalize to the same pole in Pseudomonas aeruginosa cells and to expose the regulation underlying their spatial organization and functioning.

Comments on revisions:

The authors have addressed all major and minor points that I raised in a satisfying way during the revision process. The work can now be regarded as complete, the assumptions were clarified, the results are convincing, the conclusions are justified, and the novelty has been made clear.

This manuscript will be of interest to cell biologists, mainly those studying bacteria, but not only

Reviewer #2 (Public review):

Summary:

Here, the authors studied the molecular mechanisms by which the chemoreceptor cluster and flagella motor of Pseudomonas aeruginosa (PA) are spatially organized in the cell. They argue that FlhF is involved in localizing the receptors-motor to the cell pole, and even without FlhF, the two are colocalized. Finally, the authors argue that the functional reason for this colocalization is to insulate chemotactic signaling from other signaling pathways, such as cyclic-di-GMP signaling.

Strength:

The experiments and data are high quality. It is clear that the motor and receptors co-localize, and that elevated CheY levels lead to elevated c-di-GMP.

Weakness:

The explanation for the functional importance of receptor-motor colocalization is plausible but is still not conclusively demonstrated. Colocalization might reduce CheY levels throughout the cell in order to reduce cross-talk with c-di-GMP. This would mean that if physiologically-relevant levels of CheYp near the pole were present throughout the cell, c-di-GMP levels would be elevated to a point that is problematic for the cell. Clearly demonstrating this seems challenging.

Reviewer #3 (Public review):

Summary:

The authors investigated the assembly and polar localization of the chemosensory cluster in P. aeruginosa. They discovered that a certain protein (FlhF) is required for the polar localization of the chemosensory cluster while a fully-assembled motor is necessary for the assembly of the cluster. They found that flagella and chemosensory clusters always co-localize in the cell; either at the cell pole in wild type cells or randomly-located in the cell in FlhF mutant cells. They hypothesize that this co-localization is required to keep the level of another protein (CheY-P), which controls motor switching, at low levels as the presence of high-levels of this protein (if the flagella and chemosensory clusters were not co-localized) is associated with high-levels of c-di-GMP and cell aggregations.

Strengths:

The manuscript is clearly written and straightforward. The authors applied multiple techniques to study the bacterial motility system including fluorescence light microscopy and gene editing. In general, the work enhances our understanding of the subtlety of interaction between the chemosensory cluster and the flagellar motor to regulate cell motility.

Weaknesses:

The major weakness for me in this paper is that the authors never discussed how the flagellar genes expression is controlled in P. aeruginosa. For example, in E. coli there is a transcriptional hierarchy for the flagellar genes (early, middle, and late genes, see Chilcott and Hughes, 2000). Similarly, Campylobacter and Helicobacter have a different regulatory cascade for their flagellar genes (See Lertsethtakarn, Ottemann, and Hendrixson, 2011). How does the expression of flagellar genes in P. aeruginosa compare to other species? how many classes are there for these genes? is there a hierarchy in their expression and how does this affect the results of the FliF and FliG mutants? In other words, if FliF and FliG are in class I (as in E. coli) then their absence might affect the expression of other later flagellar genes in subsequent classes (i.e., chemosensory genes). Also, in both FliF and FliG mutants no assembly intermediates of the flagellar motor are present in the cell as FliG is required for the assembly of FliF (see Hiroyuki Terashima et al. 2020, Kaplan et al. 2019, Kaplan et al. 2022). It could be argued that when the motor is not assembled then this will affect the expression of the other genes (e.g., those of the chemosensory cluster) which might play a role in the decreased level of chemosensory clusters the authors find in these mutants.

Comments on revisions:

I believe the authors have performed additional experiments that improved their manuscript and they have answered many of my comments and those of the other reviewers. I am supportive of publishing this manuscript, but I still find the following points that are not clear to me (probably I am misunderstanding some points; the authors can clarify).

(1) In response to reviewer 1, the authors say that they "analyzed and categorized the distribution of the chemotaxis complex in both wild-type and flhF mutant strains into three patterns: precise-polar, near-polar, and mid-cell localization." I can see what they mean by polar and mid-cell, but near-polar sounds a bit elusive? Can they provide examples of this stage and mention how accurately they can identify it? Also, do the pie charts they show in Figure S4 really show "significant alterations"? There is a difference between 98% and 85% as they mention in their response to reviewer 1, but I am not sure that this is significant? Probably they can explain/change the language in the text? Also, the number of cells they counted for FlhF mutant is more than the double of other strains (WT and FlhF FliF mutant)?

(2) One thing that also confused me is the following: One point that the authors stress is that FlhF localizes both the flagellum and the chemoreceptors to the pole. However, if I look at Figure 2B, the flagellum and the chemoreceptors still co-localize together (although not at the pole). If FlhF was responsible for co-localizing both of them to the pole, then wouldn't one expect them to be randomly localized in this mutant and by that I mean that they do not co-localize but that each of them (the flagellum and the chemoreceptors) are located in a different random location of the cell (not co-localized). The fact that they are still co-localized together in this mutant could also be interpreted by, for example, that FlhF localizes the flagellum to the pole and another mechanism localizes the chemoreceptors to the flagellum, hence, they still co-localize in this mutant because the chemoreceptors follow the flagellum by another mechanism to wherever it goes?

(3) In the response to reviewers, the authors mention "suggesting that the assembly of the receptor complex is likely influenced mainly by the C-ring and MS-ring structures rather than by the P ring" . However, in the article, they still write "The complete assembly of the motor serves as a partial prerequisite for the assembly of the chemotaxis complex, and its assembly site is also regulated by the polar anchor protein FlhF" despite their FlgI results which is not in accordance with this statement? Also, As I mentioned in my previous report, in FliG and FliF mutant the motor does not assemble (see Hiroyuki Terashima et al. 2020., and Kaplan et al., 2022).

(4) The authors have said in their response to my point "and currently, there is no evidence that FliA activity is influenced by proteins like FliG". I just want to clarify what I meant in my previous report: In E. coli, FliA binds to FlgM, and when the hook is assembled FlgM is secreted outside the cell allowing FliA to trigger the transcription of class III genes, which include the chemosensory genes (see Figure 5 in Beeby et al, 2020 in FEMS Microbiology, and Chilcott and Hughes, 2000). This implies that if the hook is not built, then late genes (including the chemoreceptors) should not be present. However, in Kaplan et al., 2019, the authors imaged a FliF mutant in Shewanella oneidensis (Figure S3) and still saw that chemoreceptors are present (I believe the authors must highlight this). This suggests that species such as Shewanella and Pseudomonas have a different assembly process than that E. coli, and although the authors say that in the text, I believe they still can refine this part more in the spirit of what I wrote here.

I do not like to ask for additional experiments in the second round of review, so for me if the authors modify the text to tackle these points and allow for probable alternative explanations/ highlight gaps/ modify language used for some claims, then that is fine with me.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #1 (Public review):

Summary:

The study by Wu et al presents interesting data on bacterial cell organization, a field that is progressing now, mainly due to the advances in microscopy. Based mainly on fluorescence microscopy images, the authors aim to demonstrate that the two structures that account for bacterial motility, the chemotaxis complex and the flagella, colocalize to the same pole in Pseudomonas aeruginosa cells and to expose the regulation underlying their spatial organization and functioning.

Comments on revisions:

The authors have addressed all major and minor points that I raised in a satisfying way during the revision process. The work can now be regarded as complete, the assumptions were clarified, the results are convincing, the conclusions are justified, and the novelty has been made clear.

This manuscript will be of interest to cell biologists, mainly those studying bacteria, but not only.

Reviewer #2 (Public review):

Summary:

Here, the authors studied the molecular mechanisms by which the chemoreceptor cluster and flagella motor of Pseudomonas aeruginosa (PA) are spatially organized in the cell. They argue that FlhF is involved in localizing the receptors-motor to the cell pole, and even without FlhF, the two are colocalized. Finally, the authors argue that the functional reason for this colocalization is to insulate chemotactic signaling from other signaling pathways, such as cyclic-di-GMP signaling.

Strength:

The experiments and data are high quality. It is clear that the motor and receptors co-localize, and that elevated CheY levels lead to elevated c-di-GMP.

Weakness:

The explanation for the functional importance of receptor-motor colocalization is plausible but is still not conclusively demonstrated. Colocalization might reduce CheY levels throughout the cell in order to reduce cross-talk with c-di-GMP. This would mean that if physiologically-relevant levels of CheYp near the pole were present throughout the cell, c-di-GMP levels would be elevated to a point that is problematic for the cell. Clearly demonstrating this seems challenging.

We acknowledge that directly proving the necessity of colocalization to prevent problematic c-di-GMP elevation is experimentally challenging, as it would require creating a system where CheY-P is artificially distributed throughout the cell at physiologically relevant concentrations while maintaining normal chemotaxis function.

However, our data provide several lines of evidence supporting this model. First, we show that CheY overexpression leads to substantial c-di-GMP elevation (71.8% increase) and cell aggregation, demonstrating that elevated CheY levels can indeed cause problematic cross-pathway interference. Second, previous work has shown that CheY-P levels near the pole are an order of magnitude higher than in the rest of the cell (ref. 46). If this elevated CheY-P concentration near the pole were present throughout the cell, our data suggest that c-di-GMP levels would be elevated sufficiently to cause cell aggregation (Fig. 4A), thereby disabling normal motility and chemotaxis. Third, the dose-dependent relationship between CheY concentration and aggregation phenotype supports the idea that precise spatial regulation of CheY levels is functionally important for avoiding cross-pathway interference.

Reviewer #3 (Public review):

Summary:

The authors investigated the assembly and polar localization of the chemosensory cluster in P. aeruginosa. They discovered that a certain protein (FlhF) is required for the polar localization of the chemosensory cluster while a fully-assembled motor is necessary for the assembly of the cluster. They found that flagella and chemosensory clusters always co-localize in the cell; either at the cell pole in wild type cells or randomly-located in the cell in FlhF mutant cells. They hypothesize that this co-localization is required to keep the level of another protein (CheY-P), which controls motor switching, at low levels as the presence of high-levels of this protein (if the flagella and chemosensory clusters were not co-localized) is associated with high-levels of c-di-GMP and cell aggregations.

Strengths:

The manuscript is clearly written and straightforward. The authors applied multiple techniques to study the bacterial motility system including fluorescence light microscopy and gene editing. In general, the work enhances our understanding of the subtlety of interaction between the chemosensory cluster and the flagellar motor to regulate cell motility.

Weaknesses:

The major weakness for me in this paper is that the authors never discussed how the flagellar genes expression is controlled in P. aeruginosa. For example, in E. coli there is a transcriptional hierarchy for the flagellar genes (early, middle, and late genes, see Chilcott and Hughes, 2000). Similarly, Campylobacter and Helicobacter have a different regulatory cascade for their flagellar genes (See Lertsethtakarn, Ottemann, and Hendrixson, 2011). How does the expression of flagellar genes in P. aeruginosa compare to other species? how many classes are there for these genes? is there a hierarchy in their expression and how does this affect the results of the FliF and FliG mutants? In other words, if FliF and FliG are in class I (as in E. coli) then their absence might affect the expression of other later flagellar genes in subsequent classes (i.e., chemosensory genes). Also, in both FliF and FliG mutants no assembly intermediates of the flagellar motor are present in the cell as FliG is required for the assembly of FliF (see Hiroyuki Terashima et al. 2020, Kaplan et al. 2019, Kaplan et al. 2022). It could be argued that when the motor is not assembled then this will affect the expression of the other genes (e.g., those of the chemosensory cluster) which might play a role in the decreased level of chemosensory clusters the authors find in these mutants.

We thank the reviewer for the valuable suggestions. In the revised manuscript, we have further elaborated on the regulatory control of flagellar genes expression in P. aeruginosa (see our response to comment #4).

Comments on revisions:

I believe the authors have performed additional experiments that improved their manuscript and they have answered many of my comments and those of the other reviewers. I am supportive of publishing this manuscript, but I still find the following points that are not clear to me (probably I am misunderstanding some points; the authors can clarify).

(1) In response to reviewer 1, the authors say that they "analyzed and categorized the distribution of the chemotaxis complex in both wild-type and flhF mutant strains into three patterns: precise-polar, near-polar, and mid-cell localization." I can see what they mean by polar and mid-cell, but near-polar sounds a bit elusive? Can they provide examples of this stage and mention how accurately they can identify it? Also, do the pie charts they show in Figure S4 really show "significant alterations"? There is a difference between 98% and 85% as they mention in their response to reviewer 1, but I am not sure that this is significant? Probably they can explain/change the language in the text? Also, the number of cells they counted for FlhF mutant is more than the double of other strains (WT and FlhF FliF mutant)?

We thank the reviewer for the valuable suggestions. To clarify, we divided the intracellular area along the cell's long axis into three domains: the two ends each representing 10% of the length as the precise-polar domain, the central 50% as the mid-cell domain, and the remaining regions between these as the near-polar domain. The localization pattern of the chemotaxis complex was assigned based on the position of the fluorescence intensity centroid within these domains.

Regarding the significance of the changes, you are correct to question our language. When flhF was knocked out, the proportion of chemotaxis complexes with precise-polar distribution decreased from 98% to 85% - a 13% reduction. While this represents a measurable shift in localization pattern, describing this as "significant alterations" was probably imprecise. We have revised this language to more accurately reflect the magnitude of the change (lines 169-177).

For the cell counting, we increased the sample size for the flhF mutant because this strain exhibited the appearance of mid-cell localization (approximately 5% of cells), which was not observed in wild-type or flhF fliF double mutant strains. To accurately quantify this rare phenotype and ensure statistical reliability, we analyzed more cells for this particular strain. This explains why the flhF mutant dataset contains approximately double the number of cells compared to the other strains.

We have redrawn Figure S4 to include a clear schematic diagram of the cell partitioning method and provided representative examples of each localization pattern (precise-polar, near-polar, and mid-cell) to better illustrate how we distinguished between these categories.

(2) One thing that also confused me is the following: One point that the authors stress is that FlhF localizes both the flagellum and the chemoreceptors to the pole. However, if I look at Figure 2B, the flagellum and the chemoreceptors still co-localize together (although not at the pole). If FlhF was responsible for co-localizing both of them to the pole, then wouldn't one expect them to be randomly localized in this mutant and by that I mean that they do not co-localize but that each of them (the flagellum and the chemoreceptors) are located in a different random location of the cell (not co-localized). The fact that they are still co-localized together in this mutant could also be interpreted by, for example, that FlhF localizes the flagellum to the pole and another mechanism localizes the chemoreceptors to the flagellum, hence, they still co-localize in this mutant because the chemoreceptors follow the flagellum by another mechanism to wherever it goes?

Thank you for this insightful observation. You are correct that our current experimental results do not definitively establish that FlhF directly localizes both the flagellum and chemoreceptors to the pole independently. The persistent colocalization of flagella and chemoreceptors in the DflhF mutant, even when both are mislocalized away from the pole, actually suggests a more complex regulatory mechanism than we initially proposed.

This observation highlights an important distinction between polar targeting and colocalization maintenance. Our data suggest that FlhF influences the polar targeting of the flagellum-chemoreceptor assembly, but the colocalization itself appears to be governed by a different mechanism that operates independently of FlhF. This could involve direct protein-protein interactions between flagellar and chemotaxis components, or shared assembly machinery that we have yet to identify.

To better reflect this interpretation, we have revised the subsection title (line 150). We have also modified the relevant discussion (line 180) to more accurately describe FlhF’s role in polar targeting rather than claiming it directly controls chemoreceptor localization.

(3) In the response to reviewers, the authors mention "suggesting that the assembly of the receptor complex is likely influenced mainly by the C-ring and MS-ring structures rather than by the P ring". However, in the article, they still write "The complete assembly of the motor serves as a partial prerequisite for the assembly of the chemotaxis complex, and its assembly site is also regulated by the polar anchor protein FlhF" despite their FlgI results which is not in accordance with this statement? Also, As I mentioned in my previous report, in FliG and FliF mutant the motor does not assemble (see Hiroyuki Terashima et al. 2020., and Kaplan et al., 2022).

We thank the reviewer for the suggestions and acknowledge the contradictions in our original text. You are correct that in DfliF and DfliG mutants, the flagellar motor does not assemble, while the P ring (FlgI) functions as a bushing for the peptidoglycan layer and its absence does not prevent motor assembly.

Our DflgI results, which showed normal chemotaxis complex assembly similar to wild-type, clearly demonstrate that the P ring is not required for chemoreceptor complex formation. This contradicts our original statement that "complete assembly of the motor serves as a partial prerequisite for the assembly of the chemotaxis complex."

We have corrected this inconsistency by: 1) Revising the subsection title (line 186) to more accurately reflect that core motor structures, rather than complete motor assembly, influences chemoreceptor complex formation. 2) Modifying sentences in the introduction (lines 97-98) to better align with our experimental findings.

(4) The authors have said in their response to my point "and currently, there is no evidence that FliA activity is influenced by proteins like FliG". I just want to clarify what I meant in my previous report: In E. coli, FliA binds to FlgM, and when the hook is assembled FlgM is secreted outside the cell allowing FliA to trigger the transcription of class III genes, which include the chemosensory genes (see Figure 5 in Beeby et al, 2020 in FEMS Microbiology, and Chilcott and Hughes, 2000). This implies that if the hook is not built, then late genes (including the chemoreceptors) should not be present. However, in Kaplan et al., 2019, the authors imaged a FliF mutant in Shewanella oneidensis (Figure S3) and still saw that chemoreceptors are present (I believe the authors must highlight this). This suggests that species such as Shewanella and Pseudomonas have a different assembly process than that E. coli, and although the authors say that in the text, I believe they still can refine this part more in the spirit of what I wrote here.

We thank the reviewer for the important clarification regarding the differences in transcriptional regulation among bacterial species. We agree that the observation of chemoreceptors in Shewanella oneidensis DfliF mutants (Kaplan et al., 2019) represents a significant deviation from the well-characterized E. coli model and merits stronger emphasis. In response, we have expanded the discussion to more clearly highlight the critical distinctions in the transcriptional regulatory circuits governing flagellar and chemoreceptor biogenesis between E. coli and species such as Shewanella oneidensis and Pseudomonas aeruginosa (lines 351-363).

I do not like to ask for additional experiments in the second round of review, so for me if the authors modify the text to tackle these points and allow for probable alternative explanations/ highlight gaps/ modify language used for some claims, then that is fine with me.

Reviewer #2 (Recommendations for the authors):

It is plausible that colocalization reduces CheY levels throughout the cell in order to reduce cross-talk with c-di-GMP. This would mean that if physiologically-relevant levels of CheYp near the pole were present throughout the cell, c-di-GMP levels would be elevated to a point that is problematic for the cell. Clearly demonstrating this seems challenging.

We acknowledge that directly proving the necessity of colocalization to prevent problematic c-di-GMP elevation is experimentally challenging, as it would require creating a system where CheY-P is artificially distributed throughout the cell at physiologically relevant concentrations while maintaining normal chemotaxis function.

However, our data provide several lines of evidence supporting this model. First, we show that CheY overexpression leads to substantial c-di-GMP elevation (71.8% increase) and cell aggregation, demonstrating that elevated CheY levels can indeed cause problematic cross-pathway interference. Second, previous work has shown that CheY-P levels near the pole are an order of magnitude higher than in the rest of the cell (ref. 46). If this elevated CheY-P concentration near the pole were present throughout the cell, our data suggest that c-di-GMP levels would be elevated sufficiently to cause cell aggregation (Fig. 4A), thereby disabling normal motility and chemotaxis. Third, the dose-dependent relationship between CheY concentration and aggregation phenotype supports the idea that precise spatial regulation of CheY levels is functionally important for avoiding cross-pathway interference.

Version 3

Abstract

Abstract

The opportunistic pathogen Pseudomonas aeruginosa serves as a model organism for studying multiple signal transduction pathways. The chemoreceptor cluster, a core component of the chemotaxis pathway, is assembled from hundreds of proteins. The unipolar distribution of receptor clusters has long been recognized, yet the precise mechanism governing their assembly remains elusive. Here, we directly observed the relative positions of the flagellar motor and chemoreceptor cluster using flagellar filament labeling and gene editing techniques. Surprisingly, we found that both are located at the same cell pole, with the distribution pattern controlled by the polar anchor protein FlhF. Additionally, the efficient assembly of the chemoreceptor cluster is partially dependent on the integrity of the motor structure. Furthermore, we discovered that overexpression of the chemotaxis regulatory protein CheY leads to high intracellular levels of the second messenger c-di-GMP, triggering cell aggregation. Therefore, the colocalization of the chemoreceptor cluster and flagellum in P. aeruginosa serves to avoid cross-pathway signaling interference, enabling cells to conduct various physiological activities in an orderly manner.

Introduction

In biological systems, the process by which proteins self-assemble into organized complex structures is widespread1,2. Pole-to-pole protein oscillations in the Min system ensure that the FtsZ ring, a crucial component of cell division, is placed precisely in the middle of the cell body3,4. In response to antibiotic treatment and heat stress, some bacteria generate multiple protein foci throughout the cytoplasm5,6. The polar localization of the secretion system, such as the type [secretion system, is mediated by targeting proteins and potentially facilitates host-pathogen interactions7,8. Thus, spontaneous assembly is accompanied by the formation of spatial patterns that organize the intracellular environment.

To grow and colonize in various ecological niches, bacteria have evolved a mechanism for migrating towards more favorable environments. The locomotion organ (the flagellar motor) and the chemotaxis pathway play crucial roles in achieving this goal. The former provides physical drive, while the latter offers directional guidance. The initiation of cell reorientation is controlled by motor switching, which is governed by the chemotaxis two-component signaling pathway. This pathway comprises transmembrane chemoreceptors, a histidine kinase CheA, an adaptor protein CheW, a response regulator CheY, and two adaptation proteins, CheB and CheR. Receptors collaborate with CheA and CheW to form chemotaxis complex in a hexagonal pattern9,10. Upon phosphorylation of CheY by the kinase at the complex, it freely diffuses to the motor and binds to FliM, a component of the motor switch complex, to affect motor switching11–13. It is worth noting that there exists diversity among different bacterial species in the placement of flagellar motors14, whereas chemoreceptor clusters are typically found at one or both ends of the cell15. These two structural units are functionally connected. However, it remains unclear whether there is an interaction in their spatial distribution and what the specific regulatory mechanism might be.

Pseudomonas aeruginosa, a common human opportunistic pathogen, possesses four chemosensory pathways that perform distinct functions and are stimulated by signal binding to 26 chemoreceptors16. Among them, the proteins encoded by chemotaxis-related genes collectively constitute the F6 pathway. Previous studies have suggested that the receptors involved in this pathway localize to the old cell pole17, similar to the flagellum18, implying colocalization of the receptors and the flagellum at the same pole; however, direct evidence for this colocalization is still lacking. Furthermore, the polar anchor protein FlhF has been reported to guide the flagellum to grow at the cell pole19,20. However, the mechanism by which chemotaxis complex-related proteins aggregate and self-assemble at the cell pole is controversial. The distribution of the chemotaxis complex in the peritrichous bacterium Escherichia coli has been extensively studied and attributed to several mechanisms, including stochastic self-assembly21,22, membrane curvature sorting23,24, and inefficient clustering in the lateral region25. However, these mechanisms cannot explain the unipolar distribution pattern observed in P. aeruginosa. Unlike E. coli, specific genes responsible for the placement of the chemotaxis complex have been identified in several bacterial species. In Caulobacter crescentus, chemotaxis complex assemble at the new cell pole with the help of TipN and TipF proteins26. A tripartite ParC-ParP-CheA interaction network was reported to promote polar localization of chemotaxis complex in Vibrio parahaemolyticus27,28. However, as there are no related genes in P. aeruginosa, the mechanism of its chemotaxis complex distribution remains a mystery.

Here, we combined gene editing and in vivo fluorescence imaging of flagellar filaments to directly observe the distribution of chemotaxis complex and flagellar motors, proposing a cooperative construction model of chemotaxis network and flagella during the entire division cycle of P. aeruginosa. The core focus of this study is to clarify the regulatory mechanism of its chemotaxis complex distribution. We found a substantial association between the assembly of the flagellar motor and the chemotaxis complex. The assembly efficiency of the receptor complex is influenced by core flagellar motor components, particularly the C-ring and MS-ring structures, while its assembly site is also affected by the polar anchor protein FlhF. Furthermore, by introducing exogenously expressed CheY protein, we found that this triggers the expression of c-di-GMP at a high level. From this, we infer that the colocalization of the chemotaxis complex and the flagellum in P. aeruginosa avoids the cross-pathway interference of signaling molecules, thus providing a guarantee for the coexistence of multiple chemosensory pathways.

Results

Robust generation of daughter cell with both chemotaxis network and flagellar motor

CheY has been shown to colocalize with chemoreceptors17,29. To visualize the distribution of chemotaxis complex in cells, we fused the gene encoding the enhanced yellow fluorescent protein (eYFP) to the cheY gene in the P. aeruginosa chromosome (Fig. 1A). All genetic modifications were carried out at the native position on the chromosome, allowing the expression of chemotaxis proteins in precise stoichiometry from their natural promoters (Fig. 1B). The mutant strain with cheY-eyfp was able to moderately expand on soft agar plates17, proving that its chemotaxis behavior was not notably affected, so it was selected for subsequent experiments. As shown in Fig. 1C, CheY-eYFP was mainly located at the cell pole, suggesting that CheA, CheW and CheY in P. aeruginosa form a signal transduction complex that was primarily distributed at the cell pole, as described previously17. Representative large-field images containing many cells are shown in Fig. S1. Additionally, we constructed a plasmid expressing CheA-CFP and electroporated it into the cheY-eyfp strain. Fluorescence imaging revealed a high degree of spatial overlap between CheA-CFP and CheY-EYFP (Fig. S2), confirming that CheY-EYFP accurately marks the location of the chemoreceptor complex. From our measurements, nearly 90% (332/372) of cells contain obvious receptor clusters, and factors such as fluorescence bleaching may have caused the loss of fluorescent spots in some cells. In addition, we observed the presence of fluorescent spots at both ends of some cells with a large aspect ratio (probably approaching cell division), indicating that the newly generated progeny cells will have a complete chemotaxis network. Next, we sought to observe the distribution of the flagella in cells simultaneously to understand the motility of future progeny.

A. Schematic diagram of the chemotaxis signal transduction network and flagellar motor of P. aeruginosa. Fusion protein CheY-EYFP and fluorescently labeled flagellar filaments were used as markers to indicate the position of the chemotaxis complex and motor, respectively. B. The labeling mechanism of flagellar filaments and chemotaxis regulatory protein CheY. Filaments (with cysteine point mutation FliCT394C) were labeled through sulfhydryl-maleimide conjugation, and cheY-eyfp fusion with a 3× glycine linker was used to visualize chemotaxis complex positions. C. Localization of CheY-EYFP in the wild-type strain of P. aeruginosa. CheY-EYFP is mainly located at the single cell pole, and the white arrow points to individuals with obvious chemotaxis complex at both cell poles, which generally have a large aspect ratio of the cell body. The yellow dashed box marks the cell outline. D. The merged imaging of flagellar filaments and CheY-EYFP in the wild-type strain of P. aeruginosa, where flagellar motor and chemotaxis complex colocalize in cells. 145 cells with labeled flagella were observed, all of which exhibited consistent colocalization. White arrows point to individuals about to be divided, and the yellow dashed box marks the cell outline. The scale bar is 1 μm.

Our recent study demonstrated the successful labeling of the flagellar filament with thiol-reactive fluorescent dye by introducing cysteines into the flagellin FliC30,31 (Fig. 1B). To avoid fluorescence interference, we utilized a dye with an excitation peak near 568 nm, considering that the excitation peak of eYFP is 513 nm, and the emission peak is 527 nm. We employed a xenon lamp as the excitation light source and switched filters to enable simultaneous fluorescence imaging of flagellar filaments and chemotaxis complex within the same cell.

Remarkably, our findings revealed a surprising colocalization of chemotaxis complex and flagellar filaments at the same end of the cell body, and this colocalization was consistent, as shown in Fig. 1D. A typical large field was shown in Fig. S3. Similar to the chemotaxis complex, we observed the phenomenon of double flagella symbiosis in cells with a large aspect ratio, suggesting that P. aeruginosa has evolved a cell division mechanism with precise timing regulation. Before cell division, both poles of the mother cell assembled chemotaxis complex and flagellar motors. This ensures that daughter cells possess complete motility and chemotaxis, thereby greatly enhancing the environmental adaptability of the population.

FlhF controls polar targeting but not flagellum-chemoreceptor colocalization

Despite potential differences in the physical and especially physiological environments at the two cell poles, it is unlikely that the unipolar distribution of the chemotaxis complex can be attributed to passive regulatory factors. The number and distribution of flagellar motors and chemotaxis complex vary among different bacterial species14,15, and the localization system for flagella has been well studied. Specifically, the FlhF-FlhG system, discovered in several monotrichous bacterial species, has been shown to control the location and number of flagella32,33. In P. aeruginosa, a knockout of flhF leads to mis-localized flagellar assembly19,20. Considering the consistent colocalization pattern between chemotaxis complex and flagellar motors in P. aeruginosa, we speculate that the distribution of chemotaxis complex is also regulated by similar molecular mechanisms.

To investigate the role of FlhF in the localization of chemotaxis complex, we constructed a ΔflhF strain for fluorescence observation. The results revealed that the chemotaxis complex no longer grow robustly at the cell pole (Fig. 2A), and the assembly positions of the flagellar motor change accordingly, colocalizing with the chemotaxis complex (Fig. 2B). Representative large fields containing many cells are shown in Fig. S1. Additionally, we categorized receptor cluster distribution patterns into three types: precise polar, near polar, and mid-cell. We artificially divided the intracellular area along the long axis of the cell. The two ends of the length accounted for 10% each, which we called the precise-polar domains, the middle 50% became the mid-cell domain, and the rest was called the near-polar domain. The type of complex distribution was determined by the partition where the center of the chemotactic complex fluorescent bright spot was located. The relevant statistics for the wild-type strain and the ΔflhF mutant are presented in Fig. S4, showing that the proportion of precise polar distribution of chemotactic complexes decreased by 12.8% after flhF knockout. We also quantified the proportion of individuals with obvious receptor clusters, which was more than 80% (182/221), similar to the wild-type strain. These experimental results suggest that the polar anchoring protein FlhF has a minimal impact on the assembly efficiency of P. aeruginosa’s chemotaxis complex, but it affects its unipolar distribution. Furthermore, a consistent colocalization between the flagellar motor and chemotaxis complex was observed, independent of FlhF. Thus, it is crucial to determine whether there is a causal relationship in the assembly order of these two structural units.

A. Localization of CheY-EYFP in the ΔflhF strain of P. aeruginosa. CheY-EYFP is no longer robustly distributed at the single cell pole. The yellow dashed box marks the cell outline. B. The merged imaging of flagellar filaments and CheY-EYFP in the ΔflhF strain of P. aeruginosa, flagellar motor and chemotaxis complex still colocalize in cells. 101 cells with labeled flagella were observed, all of which exhibited consistent colocalization. The yellow dashed box marks the cell outline. The scale bar is 1 μm.

Core motor structures influence chemoreceptor complex assembly

Cryo-electron microscopy has successfully revealed the complete flagella structure in various bacterial species. This structure includes multiple components such as the inner-membrane MS ring, the cytoplasmic C ring, and the internal protein secretion system, all of which exhibit strong similarities34. This suggests that the core structure of the flagellar motor is highly conserved. The MS ring, composed of 26 FliF proteins, forms the foundation of the flagellar structure35. In addition to serving as a mounting platform for the C ring, the MS ring also acts as a protective shell for the internal protein secretion machinery36. FliG, the main component protein of the C ring, binds directly to the cytoplasmic surface of FliF in a 1:1 ratio37. Thus, the efficient assembly of the C ring requires preassembly of the MS ring. To observe the distribution of the chemotaxis complex following the disruption of the C ring and MS ring, we constructed ΔfliG and ΔfliF mutants.

We performed fluorescence observation on the ΔfliG mutant, which has a disrupted flagellar motor C ring. The results showed that the proportion of ΔfliG cells with obvious chemoreceptor clusters decreased significantly compared to the wild-type cells (59.8%, 140/234), although the chemotaxis complex remained at a single cell pole. We also examined the ΔfliF mutant, which has a disrupted flagellar motor MS ring. Similar to the ΔfliG mutant, the proportion of individuals with obvious chemotaxis complex decreased significantly (62.5%, 202/323). Subsequently, we constructed a ΔflhFΔfliF mutant and observed that the proportion of individuals with obvious chemoreceptor clusters further decreased (50.7%, 341/672). To ascertain whether it is motor integrity rather than functionality that influences the efficiency of chemotaxis complex assembly, we constructed a stator mutant (ΔmotAΔmotCD). In this mutant, the motor is completely stalled while the structure remains intact. We found that the mutant performed similarly to the wild-type strain in terms of chemotaxis complex assembly (84.3% of individuals with obvious clusters, 204/242). The fluorescence imaging of receptor clusters for multiple mutants in this section is shown in Figure. 3A, and corresponding large-field images are provided in Fig. S1. We utilized Western blotting to measure the expression levels of CheY, which were found to be similar across the different strains (Fig. 3B). These further substantiated that the observed phenomenon is based on the structural integrity of the motor rather than the protein expression level. In contrast, the cheA knockout strain was constructed to disrupt the assembly of the chemoreceptor complex. We fluorescently labeled its flagellar filaments and found that its phenotype was no different from that of the wild-type strain (Fig. S5), indicating that the assembly of the chemotactic receptor complex does not affect the flagellar assembly efficiency. The P ring (mainly composed of FlgI) is thought to act as a bushing for the peptidoglycan layer, and its absence results in partial assembly of the motor structure38. We constructed ΔflgI mutant and found that the proportion of cells with distinct chemotactic complexes was similar to that of the wild-type (Fig. S6). Additionally, we quantified the proportion of cells with receptor clusters (Fig. 3C). Overall, our findings suggest that the polar anchor protein FlhF and the structural integrity of the motor are crucial for the formation of chemotaxis complex. The former guides them to the appropriate site, while the latter influences the assembly efficiency of the chemoreceptor complex.

A. Localization of CheY-EYFP in various P. aeruginosa strains. The scale bar is 10 μm. B. Western blot analysis was performed to detect CheY expression in various P. aeruginosa strains. β-actin was used as the housekeeping protein. C. Occurrence probability of obvious chemotaxis complex in P. aeruginosa wild-type and several mutant strains. The proportion of individuals with obvious chemotaxis complex decreased significantly in the motor-incomplete strains (ΔfliF and ΔfliG), and this value was further reduced after flhF knockout (ΔfliFΔflhF). The ΔfliF and ΔmotAΔmotCD strains have a similar chemotaxis complex occurrence probability as the wild-type strain. The number of cells analyzed for each strain (from left to right) were: 372, 221, 234, 323, 672, and 242. ‘***’: significant difference (P-value < 0.0001), ‘ns.’: no significant difference (P-value > 0.05).

Colocalization of chemotaxis complex and flagellar motor avoids cross-pathway interference in P. aeruginosa signal transduction

The distribution patterns of receptors associated with various signal transduction pathways in P. aeruginosa vary considerably. For instance, the biofilm formation-related receptor WspA, unlike the chemotaxis receptors discussed here, is distributed throughout the cell. This distribution is thought to enhance its sensitivity to mechanical perturbations of the cell membrane39. Previous studies speculated that the highly consistent position of chemotaxis complex and flagellar motors enhances bacterial chemotaxis performance28. However, the diffusion time of the phosphorylated chemotaxis regulatory protein CheY-P across the longest distance in a bacterial cell body (along the cell’s long axis) is approximately 100 ms40,41, whereas the timescale for the chemotaxis temporal comparison is on the order of seconds42. Additionally, a study by Fukuoka and colleagues reported that intracellular chemotaxis signal transduction requires approximately 240 ms beyond CheY or CheY-P diffusion time41. Moreover, the CCW/CW interval of the P. aeruginosa flagellar motor under normal conditions is 1-2 s, as determined by bead assays30 or tethered cell assays43. Taken together, these indicate that for P. aeruginosa, which moves via a run-reverse mode, the potential 100 ms reduction in response time due to co-localization of the chemotaxis complex and motor has a limited effect on overall chemotaxis timing. Consequently, the physiological significance of the colocalization of chemoreceptors and flagellar motors remains unclear.

E. coli mediates chemotaxis through a single pathway involving five types of chemoreceptors44. As a core regulatory protein for chemotaxis, CheY participates in multiple processes such as adaptation and chemotaxis response, with CheY-P binds to the motor to regulate switching11. In contrast, P. aeruginosa has four distinct chemosensory pathways that perform different functions, involving different CheY homologs, and are stimulated by signals binding to 26 types of chemoreceptors. Consequently, P. aeruginosa possesses more complex chemosensory pathways16. We thus hypothesized that the co-localization of chemoreceptors and flagellar motors in P. aeruginosa ensures locally distributed CheY-P molecules, thereby eliminating the need for a high level of intracellular CheY-P and avoiding potential side effects on other signaling pathways.

To test the potential effect of an increased intracellular CheY-P level, we constructed the CheY expression plasmid cheY-pJN105, transformed it into the wild-type P. aeruginosa strain, and induced it with varying concentrations of arabinose. We observed that the higher the inducer concentration, the more pronounced the cell aggregation became in the field of view (Fig. 4A). The transition from planktonic individuals to aggregated communities is similar to biofilm formation, which is known to be accompanied by a significant increase in intracellular c-di-GMP levels.

A. The evolution of cell aggregation as the intracellular CheY concentration increases by induction with higher concentrations of arabinose. The scale bar is 10 μm. B. Quantitative characterization of intracellular c-di-GMP levels at different CheY concentrations. From top to bottom, they correspond to the ΔcheY strain (N=198), wildtype strain (N=249) and CheY overexpression strain (N=228), respectively.

To investigate whether a higher level of intracellular CheY-P increased the intracellular c-di-GMP level, we sought to detect the c-di-GMP level. We introduced the plasmid pCdrA-gfp into P. aeruginosa, using it as a c-di-GMP biosensor with the fluorescence intensity proportional to the c-di-GMP level45. To ensure the coexistence of the plasmids, the intracellular CheY concentration was controlled by the plasmid cheY-pME6032. The intracellular CheY levels induced by varying arabinose concentrations were measured, showing clear differences across the concentration gradient (Fig. S7). We found that the average single-cell fluorescence intensity of the CheY overexpression strain was 71.8% higher than that of the wild-type strain, while the c-di-GMP level of the cheY deletion strain was 58.6% lower than that of the wild-type (Fig. 4B).

The colocalization of both the signaling source (the kinase) and sink (the phosphatase) at the chemoreceptor complex at the cell pole results in a rapid decay of CheY-P concentration within approximately 0.2 µm from the cell pole, leading to a nearly uniform distribution elsewhere in the cell, as demonstrated by Vaknin and Berg46. This spatial arrangement effectively confines high CheY-P levels to the pole region. When the motor is also localized at the cell pole, the need for elevated CheY-P concentrations throughout the cytoplasm is reduced. Therefore, the colocalization of chemotaxis complex and flagellar motor facilitates the precise regulation of cell swimming direction within the low intracellular CheY-P concentration threshold, using locally distributed CheY-P molecules. This helps to avoid the occurrence of the aforementioned cross-pathway interference. To further confirm that CheY overexpression promotes aggregation through increased c-di-GMP levels, we performed additional experiments co-overexpressing CheY and a phosphodiesterase (PDE) from E. coli to reduce intracellular c-di-GMP. These experiments showed that PDE expression mitigates cell aggregation caused by CheY overexpression in P. aeruginosa (Fig. S8).

Summary and Discussion

P. aeruginosa harbors multiple signal transduction systems that regulate flagella-mediated swimming motility (Che pathway), pili-mediated interface twitching motility (Pil pathway), and biofilm formation (Wsp pathway)16. These systems allow it to thrive in various ecological niches within complex external environments. Here, employing a combination of chromosomal fluorescent protein fusion and flagellar labeling techniques in living cells, we directly observed that the chemotaxis complex of P. aeruginosa Che pathway and the flagellar motor share a high degree of consistency in their assembly sites. We found that the chemotaxis complex, comprising Che proteins, was consistently located at the cell pole where the flagellum was positioned. Based on these observations, we deduced the construction mode of the chemotactic network and flagellar motor during the entire cell growth cycle (Fig. 5A). As the cell body matures and elongates, a new flagellum grows at the opposing cell pole, accompanied by the assembly of a fresh chemotaxis complex. The cell then undergoes division from the middle, generating two daughter cells, each equipped with fully functional motility and chemotaxis capabilities. This ensures robust inheritance of these chemotaxis and motility-related macromolecular machines.

A. The construction mode of the chemotaxis network and flagellar motor throughout the complete cell growth cycle. B. The proximal growth mode of the P. aeruginosa flagellar motor and receptor clusters will effectively regulate the spatial range of CheY action, thus avoiding unintentional cross-pathway regulation.

Based on our understanding of the regulation mechanism of flagellar position in P. aeruginosa, we constructed a ΔflhF mutant and found that the position distribution of its chemotaxis complex is actively regulated by FlhF-mediated molecular mechanisms, rather than being passively restricted by membrane curvature factors as seen in E. coli. A similar active regulation phenomenon was previously reported in Vibrio cholerae, with the corresponding molecular regulation mechanism detailed extensively28. The flagella and chemotaxis complex of newborn V. cholerae cells are located at the old cell pole, with the polar anchor protein HubP distributed at both cell poles. As the cell grows, HubP recruits the ParA homologue ParC. Subsequently, ParC recruits ParP to assemble a new chemotaxis complex at the new cell pole. Upon cell division, HubP relocates to the middle of the cell, ensuring its robust presences at both poles of the daughter cells. Here, HubP recruits ParA homologues including FlhG, which influence flagellar motor assembly. However, previous experiments confirmed that the cell cycle-dependent polar localization of ParC and chemotactic proteins is independent of the flagellar position regulatory protein FlhF. This suggests that the localization of flagellar motors and chemotaxis complex in V. cholerae is controlled separately through different pathways. In contrast, flagellar motors were no longer robustly distributed at the cell pole after flhF was knocked out in P. aeruginosa, whereas the assembly site of chemotaxis complex and flagellar motor remained remarkably consistent. This suggests that FlhF regulates the distribution patterns of both and implies an interaction between them.

We further constructed ΔfliG and ΔfliF mutants, which disrupted the assembly of the flagellar motor. Surprisingly, we observed that the assembly of chemotaxis complex was impaired under these conditions, with a significant decrease in the proportion of individuals exhibiting obvious Che protein fluorescent bright spots. In contrast, the chemotaxis complex assembly in the stator mutant (ΔmotAΔmotCD) was similar to that of the wild-type strain. It should be noted that P. aeruginosa has a four-tiered transcriptional regulatory circuit controlling flagellar biogenesis, in which FliG and FliF belong to class II and are regulated by the master transcriptional regulator FleQ47. Chemotaxis complex-related genes (cheA, cheW, etc.) are regulated by intracellular free FliA proteins. In E. coli, FliA is bound and inhibited by FlgM, and upon hook assembly, FlgM is secreted outside the cell, allowing FliA to trigger transcription of class III genes48,49, which include the chemosensory genes. This implies that if the hook is not assembled, late genes (including chemoreceptors) should not be expressed. However, Kaplan et al. reported that chemotaxis complexes were observed in Shewanella oneidensis ΔfliF mutant50, suggesting a different assembly process from that of E. coli. We therefore investigated whether knocking out fliF or fliG affects the expression level of che genes in P. aeruginosa. Since most experiments in this study focused on CheY, we performed Western blotting to measure CheY expression levels in wild-type, ΔfliF and ΔfliG strains, and found no significant differences (Fig. 3B). These results suggest that the observed decrease in the proportion of chemotaxis complexes after knocking out fliF and fliG should be attributed to the lack of motor assembly rather than changes in Che protein expression levels. To eliminate the potential confounding effects of FlhF and flagellar motors, we constructed a ΔflhFΔfliF mutant strain. Under this condition, the proportion of individuals with obvious Che protein fluorescent bright spots dropped to ∼50%, indicating that passive regulation based on membrane curvature may still exist in P. aeruginosa, albeit to a lesser degree.

Is the remarkable consistency in the spatial distribution of these two independent structural units an unintentional outcome or a hidden mystery? Given that Kulasekara et al. previously proposed that CheA and its phosphorylation state can cause c-di-GMP heterogeneity among individual P. aeruginosa cells through a specific PDE51,52, we sought to verify whether the phenotypic differences induced by CheY here are closely related to changes in the overall population level of c-di-GMP. We introduced a c-di-GMP monitoring system at the single-cell level and discovered that overexpression CheY substantially increases the intracellular c-di-GMP concentration. Previous studies generally believed that CheY-P, as a chemotactic regulatory protein, only influences flagellar motor switching. Here, we found that CheY-P, as a signaling molecule, can accomplish threshold-limited cross-pathway regulation. At low CheY-P concentrations, it is conservatively involved in motor switching regulation, as confirmed by previous studies. At high CheY-P concentrations, cell motility is suppressed, and c-di-GMP is triggered to express, though its specific mechanism needs further exploration. Unlike CheA, a structural protein of the chemotaxis complex, CheY is a protein that can shuttle in the cytoplasm and can also regulate the intracellular c-di-GMP level. This finding expands our understanding of the connection between the internal chemotaxis network and the c-di-GMP pathway. The intracellular environment is highly crowded53, and the proximal growth pattern shown in Fig. 5B has multiple physiological significances. On one hand, chemotaxis-related protein can precisely regulate cell motility with minimal synthesis costs. On the other hand, the chemotactic network and the flagellar motor are spatially integrated within the cell, ensuring their robust existence within the cell body and preventing mutual interference with collateral pathways.

Many bacteria possess multiple signal transduction pathways. The orderly operation of each pathway within the micrometer-scale space presents a fascinating scientific problem. Here, we utilized the well-known chemotaxis network as a starting point to shed light on this problem, providing inspiration for future in-depth understanding of related issues.

Materials and methods

Strains and cell culture

The strains and plasmids used in this study are listed in Table 1. The Escherichia coli TOP10 strain was used for standard genetic manipulations. A single-colony isolate was grown in 3 ml of LB broth (1% Bacto tryptone, 0.5% yeast extract and 1% NaCl) overnight to saturation on a rotary shaker (250 rpm) at 37°C. An aliquot was diluted 1:100 into 10 ml of LB broth and grew to exponential phase. Appropriate antibiotics were added if necessary to prevent plasmid loss: for E. coli, 15 μg/ml Gentamicin and 25 μg/ml Tetracycline; for P. aeruginosa, 30 μg/ml Gentamicin and 50 μg/ml Tetracycline. To induce protein expression, Isopropyl β-D-1-thiogalactopyranoside (IPTG, 1.0 mM) was added to strains with pME6032-derivative vectors, and arabinose (0.005%, 0.01%, 0.05%, 0.1%) was added to strains with pJN105-derivative vectors. 2 ml of cells were harvested by centrifugation at 2000×g for 2 min, washed twice in an equal volume of motility buffer (MB) [50 mM potassium phosphate, 15 μM EDTA, 0.15 M NaCl, 5 mM Mg2+ and 10 mM lactic acid (pH 7.0)]54, and resuspended in 5 ml MB for subsequent fluorescence observation.

Construction of in-frame deletion mutants

We used polymerase chain reaction (PCR) to generate ∼1000 bp DNA fragments with upstream (Up) and downstream (Dn) sequences flanking the target gene (including flhF, fliF, motA and fliG) to be deleted. The Up and Dn DNA sequences were fused with the linearized pex18gm vector using Gibson assembly55. The resulting vectors were transformed into P. aeruginosa by electroporation, and the desired knockout mutants were obtained by double selections on gentamicin plates (LB plates with 30 µg/ml gentamicin) and sucrose plates (NaCl-free LB plates with 15% sucrose) at 37°C56.

Generation of chromosomal fusions of yfp to P. aeruginosa gene

To generate eyfp fusion at the N-terminus of cheY, we first used PCR to generate upstream and downstream DNA sequences flanking the cheY gene. A subsequent PCR reaction aimed to yield the cheY-eyfp fusion with a 3× glycine linker (GGCGGAGGA), with pVS88 serving as the source of the enhanced yellow fluorescent (eyfp). The three DNA sequences, along with the linearized vector, were fused to cheY-eyfp-pex18gm using Gibson assembly. Finally, gene replacement was used to transfer the fusion constructs into the P. aeruginosa chromosome by homologous recombination.

Flagellar staining and fluorescence imaging

Flagellar filaments were labeled by following the protocol described previously57. Cells (1 ml of exponential-phase culture) were harvested by centrifugation at 2000×g for 10 min and washed twice in 1 ml MB. The final pellet was adjusted to a volume of ∼100 µl that concentrated the bacterial 10-fold. Alexa Fluor 568 maleimide (Invitrogen-Molecular Probes) was added to a final concentration of 20 μg/ml, and labeling was allowed to proceed for 30 min at room temperature with gyration at 80 rpm. Unused dye was then removed by washing the cells with MB three times, and the final pellet was resuspended in MB.

For fluorescence imaging, 50 μl of cells were added to the chamber (constructed with two double-sticky tapes, a glass slide, and a poly-L-Lysine coated coverslip), incubated for 3 min, and rinsed with 100 μl MB. The boundary of the chamber was then sealed with Apiezon vacuum grease. The chamber was placed on a Nikon Ti-E inverted fluorescence microscope with a 100× oil-immersion objective and a sCMOS camera (Primer95B, Photometrics). The flagellar filament and chemotaxis complex of cells were observed separately with corresponding filter-set, using a 200-ms exposure time. To quantify the fluorescence intensity of the receptor cluster, we took the intensity maximum point as the center and extract a 3×3 pixel matrix around it. The mean value of the elements in this matrix represents the final fluorescence intensity.

Western Blotting

The total protein of P. aeruginosa strain cells in each treatment group was extracted and its concentration was measured with Bicinchoninic acid (BCA Protein Quantification Kit, Yeasen Biotechnology Co., Ltd) when the cells grew to exponential phase with same reduced turbidity (OD600 between 0.9 and 1.0). The exact amount of protein was subjected to SDS-PAGE electrophoresis, then transferred to a nitrocellulose membrane, which was blocked with 50 mg/mL skim milk in TBST buffer (20 mM Tris, 150 mM NaCl [pH7.4], 0.1%Tween 20) for 1h at room temperature. Since the N-terminal sequence of EYFP is almost identical to that of GFP, a rabbit anti-GFP antibody (1:2000, Yeasen Biotechnology Co., Ltd) and a Mouse anti-β-actin antibody (1:2000, Invitrogen) were added and incubated at 4°C overnight. The nitrocellulose membrane was gently washed with TBST for 20 min, three times. Subsequently, HRP labeled goat anti-rabbit IgG (1:4000, Abcam) and rabbit anti-mouse IgG (1:4000, Abcam) were added and incubated at room temperature for 1 h. The nitrocellulose membrane was again gently washed with TBST for 20 min, three times. Then, the membrane was developed using the ECL chemiluminescence detection system (Beyotime P0018FS) and visualized with the ChemiDoc XRS+ system (Bio-Rad). Three independent experiments were conducted for reproducibility.

Monitoring c-di-GMP signal at the single-cell level

For measurements of c-di-GMP signaling, the validated reporter plasmid pCdrA-gfp, which produces green fluorescent protein (GFP) in response to an increase in c-di-GMP, was introduced into the experimental system. A 1000× diluted culture was inoculated into the apparatus described in the previous section and allowed to adhere to the surface for ∼10 min, with the slight difference that the coverslips were not coated with poly-L-Lysine to avoid interference from background fluorescence. Bacteria were imaged using a Nikon Ti-E inverted microscope, which was equipped with an EMCCD camera (DU897, iXon3, Andor Technology). We used a 100× oil-immersion objective. Bacteria were illuminated with a 488-nm laser (Sapphire 488−200 mW, Coherent), using a standard GFP filter sets in the fluorescence imaging system with an exposure time of 500 ms. To ensure the consistency of culture and observation conditions, the data of wild-type and cheY-pME6032 carried strains were collected at the same time. More than 200 independent individuals of each strain were used for data analysis.

Single-cell c-di-GMP concentration was characterized by calculating fluorescence intensity. The average fluorescence intensity of a single cell was calculated by dividing the total fluorescence intensity of a single cell by the cell volume. Bacterial were simplified to a hollow cylindrical trunk with hollow hemispherical caps at both ends. The cell volume can be calculated by taking the length of the short axis of the cell body as the diameter of the sphere and the length of the long axis as the sum of the height of the cylinder and the diameter of the sphere. The mean fluorescence intensity of the cell was obtained by subtracting the average background fluorescence intensity from the total fluorescence intensity of the cell, and then dividing by the cell volume.